### 2025-01-24 report/msds_data_mining_project_report.tex
add trimmed versions of sigconf acmart latex template and bibtex that are working when compiling for pdf

\documentclass[sigconf, authorversion, nonacm]{acmart}

\settopmatter{printacmref=false}

\citestyle{acmauthoryear}

\begin{document}

\title{Higher Education Synthesized}
\subtitle{Extending Higher Education Data with Synthetic Data at the Onset of the Demographic Cliff}

\renewcommand{\shortauthors}{Leakakos}

\begin{abstract}
One to two paragraph executive summary of project and presentation.

One to two paragraph executive summary of project and presentation.
\end{abstract}

\maketitle

\section{Introduction}
What is the problem?
Why is it important?
Limitations of existing solutions
What is your (potential) contribution?

\section{Related Work}
What has been done already?
Group by category
    - E.g., tools, methods, results
How does your project build upon prior work?
    - E.g., utilize, compare, improve, new knowledge

\cite{DBLP:journals/corr/abs-1907-00503}

\cite{decristofaro2024syntheticdatamethodsuse}

\section{Proposed Work}
Dataset, tools
Main tasks
    - E.g., statistical analysis, visualization
    - E.g., data preprocessing, warehousing
    - E.g., specific questions, patterns to explore or model

\section{Evaluation}
Evaluation metrics
    - Effectiveness
    - Efficiency
Experimental setup
Methods to compare

\section{Discussion}
Project timeline
    - When to finish
    - Current status
    - Determine if you need to do less or can do more in the given time
Potential challenges
Alternative approaches and backup plan

\section{Conclusion}
Conclusion
Project summary
Key findings (when finished)
Future work (when finished)

\bibliographystyle{ACM-Reference-Format}
\bibliography{msds_data_mining_project_report}

\end{document}
\endinput

### 2025-01-31 report/msds_data_mining_project_report.tex
add first pass of abstract and introduction along with updated bibliography

@@
 \begin{abstract}
-One to two paragraph executive summary of project and presentation.
+    American higher education is coming to an impasse.
+Colleges and universities have seen enrollments grow as larger generations of Americans made their way through postsecondary education.
+Schools increased their academic offerings and student life activities.
+Faculty and staff members filled the needs for supporting increasing student enrollments.
+But there was an issue hiding behind all of this growth: what happens when student enrollments systematically fall?
 
-One to two paragraph executive summary of project and presentation.
+    The United States has seen declines in high school graduates and, therefore, reduced college applicants.
+Depending on who you listen to, this ranges from a demographic cliff ushering in an emergency in higher education to a slower decline in higher education enrollments that will play out over years.
+It is unclear what effect this will have on the near future of higher education.
+One expected consequence of these declines is that we will see smaller amounts of higher education data relative to previous years.
+
+    This project focuses on synthesizing Integrated Postsecondary Education Data System data from the National Center for Education Statistics.
+This data covers a vast range of information about higher education, and, while we will not look at student-level data --- one of the key areas for synthetic data after the demographic cliff hits --- we will start by generating synthetic data based on subsets of institution-related data.
+The goal is to show how we might address the diminishing data after the demographic cliff in order to continue higher education data analysis that relies on larger datasets.
 \end{abstract}
@@
 \section{Introduction}
-What is the problem?
-Why is it important?
-Limitations of existing solutions
-What is your (potential) contribution?
+
+    American higher education saw growing enrollments throughout the latter half of the 1900s and moving into the 200s.
+This resulted in increases in the number of higher education institutions (HEIs) in the country, including public and private colleges and universities.
+These HEIs span from major research institutions down to small liberal arts schools.
+This has also resulted in increases in the number of faculty and staff to support growing enrollments.
+The general trend in higher education was growth.
+
+    The issue is that this growth in enrollments was not to last.
+For years now, we have seen predicted and then realized declines in high school graduates along with the respective drop in higher education enrollments.
+Higher education's future was based on growth, but then the key aspect of that growth --- more students than in previous years --- fell away.
+Estimates for what comes next vary.
+Some see a demographic cliff and a looming emergency in which declining enrollments result in catastrophic cuts at and closures of HEIs.
+It is true that we have seen increasing numbers of HEI closures in recent years ~\cite{hechingercollegeclosures}.
+But others see the trend as a more gradual decline, giving higher education time to adapt and shrink accordingly.
+\cite{insidehighereddemographiccliff}
+
+    There are many interesting research questions related to the future of higher education, appropriate and responsible responses to enrollment declines, and other aspects of this impending reality.
+There are also questions related to the shifting of costs from public funding to individual students, the shift in emphasis to higher education as workforce preparation, the modern publishing culture in higher education and how that affects the student experience, and more.
+
+    But, this project will instead focus on a different aspect of declining higher education enrollments: reduced data.
+As student enrollments drop and HEIs close or shrink, we are likely to see reduced amounts of higher education data available for analysis, prediction, evaluation, and more.
+Not all analyses in higher education need larger amounts of data, but we want to explore how we could supplement real data with synthetic data in cases where we need more overall data.
+The risk is that, without sufficient data, we may not be able to support students and HEIs as well as we could with larger amounts of data.
+
+    We will take a look at Integrated Postsecondary Education Data System (IPEDS) data from the National Center for Education Statistics (NCES), housed in the U.S.
+Department of Education's Institute of Education Sciences (IES).
+NCES gathers, prepares, and makes available data along with providing analysis on education data.
+IPEDS is a subset of NCES data that focuses on postsecondary or higher education data.
+
+    We will focus on a subset of IPEDS data related to institutions, enrollments, and graduation rates.
+For each of these three domains, IPEDS offers multiple files covering different information, and it provides these for multiple years.
+Our goal will be to identify one or more analysis or modeling datasets that we can pull out of a larger extract-transform-load (ETL) pipeline with multiple repeated steps.
+We will focus on pulling IPEDS data, loading it into a relational database in its raw form, preprocessing it in the database with a series of automated processes, storing the analysis or modeling datasets in the database, and then using those datasets to synthesize new data.
+
+   To address the issue of the amount of data available to us, we will leverage generative modeling to produce synthetic tabular data.
+The goal is to build a system that can demonstrate how we might leverage synthetic higher education data to fill in datasets in the future as they potentially shrink due to the issues described above.
+
+    We are focusing on institution-level data from IPEDS, but future iterations of this work could pivot to generating synthetic student-level data.
+
+    Synthetic data is ``artificially generated [data] that resemble[s] the actual data -- more precisely, having similar statistical properties'' ~\cite[p.
+1]{decristofaro2024syntheticdatamethodsuse}.
+If we generate synthetic IPEDS data that shares the same properties as the real data, then we can combine the real and synthetic data, or, alternatively, we can create an indefinite amount of synthetic data, to increase the amount of data we have to work with.
+
+    Building and training models to generate synthetic tabular data -- tabular meaning data in table form with rows and columns -- is notoriously difficult.
+We will leverage existing models and research for the synthetic generation aspect of the pipeline in particular.
+We will lean on the research in the Modeling Tabular Data Using Conditional GAN paper ~\cite{DBLP:journals/corr/abs-1907-00503} and the associated Python libraries they make available at the Synthetic Data Vault Project ~\cite{sdv}.
+
+    While we will not dive too deep into the working of the generative models, it can be useful to understand the complexity in these and why we are leveraging existing implementations.
+The research paper focuses on their conditional tabular generative adversarial network (CTGAN) approach to generative modeling.
+This paper provides some examples of the difficulty of modeling for tabular data.
+Section 3 speaks to some of these issues:
+
+    \begin{itemize}
+        \item Mixed data types: tabular data can have features with continuous, nominal, ordinal, binary, or other data.
+Generative modeling for these datasets require handling mixtures of data types in one model.
+        \item Non-Gaussian distributions: continuous data can be non-Gaussian, so attempts to standardize or min-max continuous features can result in issues such as vanishing or exploding gradients.
+        \item Multimodal distributions: continuous features may have more than one mode, resulting in the need to model more than one mode with approaches such as kernel density estimation and Gaussian mixture models.
+        \item Imbalance in categories for categorical features: categories that have smaller representation may not be modeled as accurately as those with larger representation, resulting in the need for extra steps to handle this disparity.
+\cite{DBLP:journals/corr/abs-1907-00503}
+    \end{itemize}
+
+    We do not need to go into the inner workings of these issues.
+Instead, they are meant to provide a snippet of the difficulties we would need to handle if building our own generative models.
+In order to remain focused on the pipeline in this project, we will, as said above, leverage existing generative modeling libraries available through the Synthetic Data Vault Project ~\cite{sdv}.
+
+    To summarize, this project will work with IPEDS data with its focus on higher education and will explore how we can synthesize more IPEDS data to address potential future lack-of-data issues due to declining higher education enrollments.
 
@@
 \cite{decristofaro2024syntheticdatamethodsuse}
+
+\cite{sdv}
+
+\cite{ipeds}
 
@@
 
+\appendix
+
+\section{Technical Details}
+
+    Need some details plz
+
 \end{document}

### 2025-02-07 report/msds_data_mining_project_report.tex
fill out first pass on related work section and associated bibliography entries

@@
 
-\settopmatter{printacmref=false}
+\settopmatter{printacmref=false} % remove acm info that clutters things up for this project
+\settopmatter{printfolios=true} % add page numbers
 
@@
 
-    To summarize, this project will work with IPEDS data with its focus on higher education and will explore how we can synthesize more IPEDS data to address potential future lack-of-data issues due to declining higher education enrollments.
+    Finally, the ultimate goal is to glean insights from the IPEDS data, real or synthesized.
+The synthetic data will ideally give us the same overall insights as the real data.
+For this task, we can turn to traditional data mining techniques: classification, clustering, association mining, rule mining, or others.
+If we can determine trends now, we may be able to utilize synthetic data in the future with dwindling real data to confirm continued trends or to identify changes in those trends.
+
+    To summarize, this project will work with IPEDS data with its focus on higher education and will explore how we can synthesize more IPEDS data to address potential future lack-of-data issues due to declining higher education enrollments, and we will use these data to tackle standard data mining questions related but focused on higher education data.
 
 \section{Related Work}
-What has been done already?
-Group by category
-    - E.g., tools, methods, results
-How does your project build upon prior work?
-    - E.g., utilize, compare, improve, new knowledge
+    My literature review did not return as much related to data mining in higher education as I expected to see, and I was unable to find any research closely related to what we explore in this project.
+The bulk of what I found is related to different evaluation methods for student learning.
+Examples are looking at adaptive learning --- learning that dynamically adjusts to students based on different input metrics --- and performance prediction --- trying to determine how a student will end up performing by, for example, graduation.
+While these are interesting, they do not focus on the type of data that we see in IPEDS.
 
-\cite{DBLP:journals/corr/abs-1907-00503}
+    Note that some of these papers were already referenced in the introduction above.
 
-\cite{decristofaro2024syntheticdatamethodsuse}
+    To start with, ~\cite{sdp} by the Harvard Strategic Data Project (SDP) fellowship program speaking about the importance of analytics in higher education data.
+There are larger questions that this paper raises about why higher education appears behind other areas in terms of leveraging data and more advanced analytics.
 
-\cite{sdv}
+    Another example of an interesting paper that I want to highlight is ~\cite{Hassna02102023} that explores utilizing big data and analytics for higher education data.
+The catch is that this paper is from 2022, and it is talking about introducing big data and analytics in higher education, mirroring some of what we see from the SDP paper above in hinting that higher education may be behind other areas in terms of utilizing data.
+But, during the literature review, I checked out big data and data mining and related terms in other domains, and I did not tend to see papers talking about how big data and analytics have yet to be realized in those domains.
 
-\cite{ipeds}
+    Next is a handful of papers characterizing what I mostly found: research into specific uses of analytics and mining to hone in on an aspect of student evaluation.
+Since our focus on this project is on institution-level information with IPEDS data, we are zooming out quite a bit from what these papers cover, but these are useful in terms of getting bearings for research on higher education data mining, though none of these papers are overly useful for the specific use case in this project.
+
+    First is ~\cite{studentbehavioranalytics} that talks about using higher education big data to evaluate student behavior.
+This paper repeats a similar claim to ~\cite{Hassna02102023} above in that higher education is missing work in big data.
+
+    \cite{Okewu10112021} explores using artificial neural networks for higher education data mining.
+It provides a review of studies from 2010 through 2018 and related to education data mining focused on adaptive learning.
+Adaptive learning is the use of learner data to customize learning dynamically to those learners.
+So, we are looking at yet another narrower use case for a specific type of higher education learner data.
+This paper again mirrors some of the sentiments from earlier papers in that it hints that higher education data usage, and big data usage in particular, should be further along than where it is currently at (as of the paper's publishing in 2021), closer to where we see big data usage in other domains, but it is not.
+
+    \cite{10.3389/fpsyg.2021.698490} investigates education data mining techniques for student performance prediction.
+There is not a ton more to add here aside from it being another narrower use case looking at student-level data, not the larger institution-level data that we explore in this project, but it is yet another example to learn from in terms of research using higher education data.
+
+    \cite{ALDOWAH201913} digs into education data mining, similar to the last paper, but this time focused on learning analytics.
+It reviews studies in different domains related to learning analytics.
+The key takeaway is that it is yet another study that looks into different outcomes within specific areas of higher education data, but it is not close enough to what we are exploring in this project to lean on.
+
+    Finally, ~\cite{https://doi.org/10.1155/2022/8924028} looks at predicting higher education student performance, typically at graduation, and it evaluates different predictive methods.
+Again, this is interesting for reviewing some current literature, but it will not apply to the work we do in this project.
+
+    We will also include a quick review of a few papers related to synthetic data and synthetic tabular data generative modeling.
+We will utilize synthetic data, but we are not building it from scratch, so we will not dwell on the deep learning implementation details of these papers.
+
+    \cite{decristofaro2024syntheticdatamethodsuse} provides an overview of synthetic data that is helpful for understanding the need for, uses of, and issues with synthetic data.
+We will use a production library for synthetic modeling, so we will not address these concerns directly except as parameters to the model(s) we decide to use in the pipeline, but it is good to understand the kinds of trade-offs that are inherent in generative modeling for synthetic tabular data.
+
+    \cite{DBLP:journals/corr/abs-1811-11264} introduces table GAN (TGAN), a GAN-based approach to generating synthetic tabular data.
+This model uses recurrent neural networks.
+Each of the generative modeling papers I reviewed have extensive coverage of how to handle concerns in the data.
+One interesting one to highlight, among a few, from this paper and demonstrating the type of complexity that goes into this type of modeling is the focus on multimodal continuous data.
+This is continuous data that has more than one value that stands out as more represented in the data, showing up as more than one peak when you plot a histogram of the data.
+To get around this, the researchers use Gaussian kernel density estimation (KDE) to estimate the distribution of the data.
+Each KDE can model one mode of the data.
+We then use Gaussian mixture methods (GMMs) to sample from more than one distribution -- each distribution coming from one of those built with the KDEs -- to recreate the actual distribution.
+We then sample from this GMM when we are generating a value for that feature, and this gets baked into the modeling, training, and generation processes.
+This is just one example to demonstrate why we are veering away from building our own generative models in this project.
+
+    \cite{DBLP:journals/corr/abs-1907-00503}, by the same authors as the previous TGAN paper, provides more consideration for nuances and potential issues in the data along with adding a deep conditional GAN (DCGAN) as part of their new conditional tabular GAN (CTGAN) model.
+DCGAN generally is a GAN architecture that uses convolutional neural networks (CNNs) in both the generator and discriminator.
+One example of extra data considerations is that CTGAN handles severe imbalance in categorical features.
+We will not worry about the detail of this model either, but it is potentially one that we will use with the modeling library we introduce next
+
+    \cite{sdv} is our project of choice for generative modeling.
+It is produced by the authors of the previous two papers above.
+Similar to those two papers, we will not dig too far into the workings of the libraries in this project.
+We mostly need a solution that can plug into the synthetic tabular data generative modeling step(s) in the pipeline.
+Note that both TGAN and CTGAN are dedicated repositories in the Synthetic Data Vault project.
+There is another main library -- sdv -- that provides easier-to-use versions of the models.
+There is also SDGym, a synthetic data evaluation library that we can lean on to determine how well our generative modeling is doing.
+
+    With a sweep through some research papers out of the way, I did not find any papers directly applicable to what we are looking at in this project.
+So, the contribution that we provide is to start a discussion around data mining for IPEDS data along with generative modeling for IPEDS data.
+We build on the ideas in the papers above -- the general approaches to analysis in the earlier papers, and the use of synthetic tabular data for the latter papers.
 
@@
 
+\subsection{Questions for the Reader}
+    \begin{itemize}
+        \item Is adding in the synthetic data component too much?
+I've been exploring this is personal and school projects as well as using it at work, so it's fresh on my mind to keep testing out, but it may be overkill for this project.
+    \end{itemize}
+
 \section{Conclusion}

### 2025-02-08 report/msds_data_mining_project_report.tex
add first pass of proposed work section

@@
 \section{Proposed Work}
-Dataset, tools
-Main tasks
-    - E.g., statistical analysis, visualization
-    - E.g., data preprocessing, warehousing
-    - E.g., specific questions, patterns to explore or model
+
+    For me, at this stage, this is the section that I am looking to narrow down a bit (and I am open to suggestions from readers).
+I have a couple of ideas planned for the work in this project, and I am looking to lean into my background with the architecting of data systems to build out a wider pipeline.
+
+    We are exploring IPEDS data.
+We went over the value of that data earlier.
+To snag the data, we will use some Python scripts to download the data from the IPEDS website and store it in a relational database.
+This database will run locally in Docker, and the configuration files needed to set up the database will exist in this repository, both for the Docker setup as well as for the database work.
+IPEDS data already comes in a relatively relational form, so there may not be too much work for the initial load into the database.
+
+    From there, we may leverage the database further to provide preprocessing in the database.
+One easy way to do this is to set up insert triggers to handle the first pass of standardizing the data into a consistent format or doing some early combining and mutating of data before looking at larger exploratory data analysis (EDA).
+This may work, but we may also want to pull the data out of the database to go through a more standard EDA.
+But, there is an option here to mimic what we may need in a larger production pipeline with leveraging database tech at this stage.
+
+    Next we go through a full EDA process, this time with a mind towards how we may want to use the data later in the pipeline rather than the earlier EDA based around simple cleanup.
+This step will happen outside of the database, so we will look at Python scripts here.
+At first, this is likely easier with the read-eval-print loop (REPL) functionality of a notebook, though we do not want notebooks in our final pipeline until much closer to the end in production.
+We need the automation power in scripts instead of notebooks.
+
+    After this round of EDA and any further preprocessing we glean from it, we can then think about warehousing.
+We will shift the data back into the warehouse, likely just matching the new form of the data to what we need for modeling, but we can decide later on the form of the data in the warehouse.
+For the scope of this project, we will likely end up with a bespoke warehouse built to immediate modeling or analysis needs rather than a broader warehouse set up in something like a full star schema.
+
+    The next step is to explore the outputs of the pipeline.
+Two directions stand out to me.
+One is modeling.
+Especially when working with larger datasets such as those we deal with in data mining, prediction and clustering jump out as obvious candidates, with clustering typically being much more complicated to set up.
+The second direction is to explore some data mining-specific approaches and algorithms such as similarity measures, apriori, FP-growth, association or rule mining, or others.
+Honestly, at this stage, any of these could be valid, and I prefer to start working with the data some before deciding on outputs at this stage, so this is left purposely broad.
+
+    One piece I am pulling to the end here is the synthetic data component.
+I am already wondering if adding in synthetic data is too much.
+I have been exploring synthetic data in personal and school projects as well as using it at work, so it is fresh on my mind, and I would like to keep testing it out, but it may be overkill for this project.
+If we do keep synthetic data, then it makes sense to simplify some of the rest of the pipeline to keep more of the scope focused on the synthetic data modeling.
+This means we probably focus more at the end of the pipeline on evaluating the quality of the synthetic data and demoing how we would pull insights out of the synthetic dataset, but we probably look at just one approach for that last piece instead of going through a heavy refinement of the other outputs.
+
+    The areas I am keeping an eye on here are the extra loop of preprocessing in the database, the inclusion of synthetic data, and narrowing down the final outputs.
+The ideas in this section are fine for now, but we need to make these more concrete as we begin digging into the data.
 
 \section{Evaluation}
+
+    Tests for synthetic data
+    data mining algorithms
+
 Evaluation metrics
@@
 Alternative approaches and backup plan
-
-\subsection{Questions for the Reader}
-    \begin{itemize}
-        \item Is adding in the synthetic data component too much?
-I've been exploring this is personal and school projects as well as using it at work, so it's fresh on my mind to keep testing out, but it may be overkill for this project.
-    \end{itemize}
 

### 2025-02-08 report/msds_data_mining_project_report.tex
finish first pass of project report, before starting technical work

@@
 \section{Introduction}
-
     American higher education saw growing enrollments throughout the latter half of the 1900s and moving into the 200s.
@@
 One example of extra data considerations is that CTGAN handles severe imbalance in categorical features.
-We will not worry about the detail of this model either, but it is potentially one that we will use with the modeling library we introduce next
+We will not worry about the detail of this model either, but it is potentially one that we will use with the modeling library we introduce next.
 
@@
 \section{Proposed Work}
-
     For me, at this stage, this is the section that I am looking to narrow down a bit (and I am open to suggestions from readers).
@@
 
+    I have spent most of the time here focusing on the technical side of the work.
+I need to dig into the IPEDS data before feeling more comfortable with the questions in terms of meaning that we will explore, but I expect those to center around identifying patterns in clusters of institutions based on institution attributes, enrollment trends, and graduation trends, possibly even dropping either enrollment or graduation data in favor of simplifying the data scope for right now.
+
 \section{Evaluation}
+    Until we decide to drop synthetic data, if we do make that decision, the first evaluations we will explore are how similar the synthetic data is to the real data.
+We will explore the evaluation metrics in the Synthetic Data Vault Project.
+A couple of candidates that stand out are per-attribute comparisons and machine learning score.
+Per-attribute comparisons check the same feature in the real and synthetic datasets, calculate metrics such as mean or standard deviation, and determine the quality of the synthetic data based on how close these measures are across the datasets.
 
-    Tests for synthetic data
-    data mining algorithms
+    Machine learning score trains classifiers on the real and synthetic datasets and compares the performance of those models.
+For example, if we use both classifiers to check something like accuracy or F1 score on a held-out test set, we can evaluate how similar the synthetic data is to the real data based on comparing those scores.
 
-Evaluation metrics
-    - Effectiveness
-    - Efficiency
-Experimental setup
-Methods to compare
+    We can also look at more advanced evaluation metrics.
+One that has stood out to me predicting each feature in a dataset with all of the other features.
+If we find the predictive power in the real and synthetic datasets are similar, then, like above, that can give us a sense of how similar the datasets themselves are.
+
+    There are any number of other evaluative metrics for synthetic data.
+We will explore what is available with the project we use for synthetic data and will clarify here as we learn more during experimentation.
+
+    Moving on from synthetic data, until we explore the IPEDS data further, I am not sure exactly what we will end up modeling or testing as output.
+But there are some general contenders that we could likely look at.
+
+    For supervised modeling outputs, we can look at standard metrics such as accuracy, precision, F1 score, and more for classification problems.
+We can use a metric like mean squared error (MSE) to measure predictive model output.
+These evaluations give us firm numbers we can use to determine how well our modeling is doing.
+
+    But, I am actually more intrigued by seeing what we can find with unsupervised learning.
+Admittedly, the outputs for unsupervised learning can be more difficult to work with, especially at this early stage.
+Clustering is the first unsupervised modeling that comes to mind.
+I often think of clustering as a much more advanced version of EDA.
+We are not really sure what patterns exist in the data, so we supercharge standard EDA and see what patterns we uncover when leveraging machine learning.
+But we do not have a clean evaluation metric with clustering.
+We will need to interactively explore the clustering outputs, at least at first.
+At some point, we can determine a set number of clusters and pivot the clustering into a classification approach, but the more interesting piece here for me is the clustering itself.
+
+    I am on the fence about leaning more into the synthetic data approach in which we would spend a lot more time evaluating the quality of the synthetic data.
+With synthetic data, I still would like to test out one of the analytical pieces at the end of the pipeline, but maybe we stick to something simpler like a clearer classification problem.
+If we instead drop synthetic data, then we have more room to tackle something like unsupervised learning and the complexity inherent in it.
+
+    While I feel like there are loads of evaluative methods to explore, I am also interested in looking at some of the algorithms from the second data mining course.
+I have leveraged things like similarity scores in other projects, but I have not experimented with, for instance, FP growth, and it would be interesting to possibly pick a brand new algorithm to plop at the end of the pipeline and figure out what it means to evaluate that algorithm and what it reveals to us.
+
+    So, at this stage, this section is more of a brainstorming of some of the evaluative approaches.
+This will get refined down in the next iteration after exploring the IPEDS data a bit and determining if we will keep or drop the synthetic data component of the pipeline.
 
 \section{Discussion}
-Project timeline
-    - When to finish
-    - Current status
-    - Determine if you need to do less or can do more in the given time
-Potential challenges
-Alternative approaches and backup plan
+    I tend to drag out projects, taking the opportunity to add increased complexity and explore new items I am not as familiar with.
+For this project, part of me wants to focus on sticking a bit more in-bounds in terms of my comfort level with data pipelines, only adding in a little extra complexity and seeing if it is feasible to build the entire pipeline out in a shorter amount of time.
+While spending forever on projects is fun and educational, I do risk getting too comfortable with taking my time.
+I could benefit from descoping and focusing on getting smaller chunks working earlier and then deciding if it is worth expanding out from there.
+
+    In that vein, I am hoping that the refinement in going from the first to second draft of this report, spending some time to explore the technical and data sides of the project in between, will help reveal scope from this draft that can be cut.
+It would be ideal if I could use a three or four week timeline and adjust the project to fit within that, but it does mean that what I have laid out above is probably too ambitious for that timelines.
+But, after writing this all out, the scope does feel pretty big at the moment.
+
+    Currently I have identified some IPEDS data, but I have not written any code or set up the database.
+My approach will be to get the data through to at least the initial EDA and preprocessing before starting to evaluate the scope after that.
+I feel comfortable with the earlier and more technical steps in terms of time, so my question marks sit with the longer EDA, the modeling, and, if included, the synthetic data portion.
+
+    The main challenge I see is that the data ends up being much more complicated to work with than expected, so I will need to shift the scope down even further to prioritize working with difficult data.
+That would not be a bad way for the project to pivot, but it is not what I am planning on at this stage.
+If the data turns out to be friendly enough to work with, then I think the next big challenge will be figuring out the right analysis or modeling in terms of the meaning of the data.
+The ideas above are all interesting from a technical standpoint, but the meaning of what we dig into needs to drive why we are using those approaches.
+
+    The general backup plan is to limit each stage of the pipeline to maybe one or two pieces.
+This may mean we do some simple EDA along with a correlation matrix, we look at a clearer regression or classification problem, or we leverage synthetic data but trim the input dataset to only a few features.
+All of these are tactics for trimming the scope, likely the biggest risk for the project at the moment.
 
 \section{Conclusion}
-Conclusion
-Project summary
-Key findings (when finished)
-Future work (when finished)
+    Not much to conclude at the moment.
+We will fill this in as we start doing the data work of the project.
+What I expect to see here is that we trimmed down the IPEDS data, scoped the use case for the project to be a bit narrower, decided to explore synthetic data, and picked a simpler overall output.
+For a surprise finish, we may end up deciding to pull some data out and run it through a separate business intelligence tool like Tableau with the idea that end users of the production pipeline may want to leverage tools such as that for their purposes.
+
+    I expect the next steps to end up being that we incorporate more IPEDS data to answer a broader range of questions.
+I do not expect radical changes in the pipeline architecture in next iterations of the project, but I do expect additions and expansions.
 
@@
 \section{Technical Details}
-
-    Need some details plz
+    Need some details plz. Leaving this as a placeholder to pull the technical details out of the narrative above but still keep it available for readers.
 

### 2025-02-17 artifacts/report/msds_data_mining_project_report.tex
update report while working on slides, add slides, and adjust gitignore for refactoring directory structure

--- 2025-02-07 report/msds_data_mining_project_report.tex
@@
 
-\citestyle{acmauthoryear}
+%\citestyle{acmauthoryear}
 
@@
 \subtitle{Extending Higher Education Data with Synthetic Data at the Onset of the Demographic Cliff}
-
-\renewcommand{\shortauthors}{Leakakos}
 
@@
 \section{Introduction}
-
-    American higher education saw growing enrollments throughout the latter half of the 1900s and moving into the 200s.
+    American higher education saw growing enrollments throughout the latter half of the 1900s and moving into the 2000s.
 This resulted in increases in the number of higher education institutions (HEIs) in the country, including public and private colleges and universities.
@@
 IPEDS is a subset of NCES data that focuses on postsecondary or higher education data.
+\cite{ipeds}
 
@@
     We will also include a quick review of a few papers related to synthetic data and synthetic tabular data generative modeling.
-We will utilize synthetic data, but we are not building it from scratch, so we will not dwell on the deep learning implementation details of these papers.
+We will utilize synthetic data, but we are not building generative models from scratch, so we will not dwell on the deep learning implementation details of these papers.
 
@@
 So, the contribution that we provide is to start a discussion around data mining for IPEDS data along with generative modeling for IPEDS data.
-We build on the ideas in the papers above -- the general approaches to analysis in the earlier papers, and the use of synthetic tabular data for the latter papers.
+We build on the ideas in the papers above -- the general approaches to analysis in the earlier papers, and the use of synthetic tabular data as described in the latter papers.
 
 \section{Proposed Work}
-Dataset, tools
-Main tasks
-    - E.g., statistical analysis, visualization
-    - E.g., data preprocessing, warehousing
-    - E.g., specific questions, patterns to explore or model
+
+    The general approach for the work in this project is to leverage Python scripts to move information around, a relational database running in Docker for any pieces of the pipeline that would normally use a database, and notebooks as needed for exploratory and visualization purposes.
+
+    To snag the IPEDS data, we will use Python scripts to download the data from the IPEDS website and store it in a relational database, possibly storing the data on disk as a stepping stone.
+The database will run locally in Docker.
+IPEDS data already comes in a relatively relational form, so there may not be too much work for the initial load into the database.
+Much of the data we end up working with in the real world lives in a database, so this recreates that reality here.
+
+    We may leverage the database for preprocessing.
+We would need to do some manual exploration first in Python, but, after that, we could look at shifting preprocessing into structured query language (SQL), particularly stored as insert triggers in the database.
+Right now, there is a question about how that would intersect with exploratory data analysis (EDA) that we would end up doing in Python, but we can figure out the specifics as we dig into the data.
+My thought here is that a production pipeline would be simpler if we could handle more of the preprocessing automatically in the database and not require that someone hook into separate Python scripts, whether automated or needing to be run manually.
+
+    We will need to go through a full EDA process.
+This step will happen outside of the database, so we will look at Python scripts here.
+At first, this is likely easier with the read-eval-print loop (REPL) functionality of a notebook, though we do not want notebooks in our final pipeline until much closer to the end in production.
+We need the automation power in scripts instead of notebooks.
+We may also learn more from the EDA that we can bake into extra preprocessing, informing what is mentioned in the last paragraph.
+
+    After this round of EDA and any further preprocessing we glean from it, we can then think about warehousing.
+We will shift the data back into the database, likely just matching the new form of the data to what we need for modeling, but we can decide later on regarding the form of the data in the warehouse.
+For the scope of this project, we will likely end up with a bespoke warehouse built to immediate modeling or analysis needs rather than a broader warehouse set up in something like a full star schema.
+
+    The next step is to explore the outputs of the pipeline.
+It occurs to me that there are two competing outputs in this project: synthetic IPEDS data and standard evaluative findings.
+We will need to find a balance here, so we may limit the amount of regression, classification, or clustering we do to leave space for the synthetic data generative modeling, possibly demoing a smaller example of how we would glean insights out of the synthetic dataset using classification, for example, rather than adding more scope into that outputs after the generative modeling.
+The final outputs I have a sense will work well with the synthetic IPEDS data are regression for number of enrollments or degree completions, clustering to find groupings of institutions that we should keep an eye on, and classification for putting institutions into those groupings.
+
+    I have been exploring synthetic data in personal and school projects as well as using it at work, so it is fresh on my mind, and I would really like to include it here, but, as said above, it is in tension with what we might see in a standard data mining pipeline.
+We will need to pull data out of the warehouse, import the synthetic data libraries, run the generative modeling, and then possibly load the synthetic data back into the database.
+This area has a bit of a question mark as to how much scope it will take up.
+I feel comfortable with how the generative modeling will work, but we will need to explore performance with IPEDS data in particular.
+
+    The areas I am keeping an eye on here are the extra loop of preprocessing in the database, the inclusion of synthetic data, and narrowing down the final outputs.
+The ideas in this section are fine for now, but we need to make these more concrete as we begin digging into the data.
 
 \section{Evaluation}
-Evaluation metrics
-    - Effectiveness
-    - Efficiency
-Experimental setup
-Methods to compare
+    The first evaluations will capture how similar the synthetic data is to the real data.
+We will explore the evaluation metrics in the Synthetic Data Vault Project, and, on top of those metrics, a couple of extra candidates that stand out are per-attribute comparisons and machine learning score.
+Per-attribute comparisons check the same feature in the real and synthetic datasets, calculate metrics such as mean or standard deviation, and determine the quality of the synthetic data based on how close these measures are across the datasets.
+Machine learning score trains classifiers on the real and synthetic datasets and compares the performance of those models.
+For example, if we use both classifiers to check something like accuracy or F1 score on a held-out test set, we can evaluate how similar the synthetic data is to the real data based on comparing those scores.
+
+    We can also look at more advanced evaluation metrics.
+One that has stood out to me predicting each feature in a dataset with all of the other features.
+If we find the predictive power in the real and synthetic datasets are similar, then, like above, that can give us a sense of how similar the datasets themselves are.
+
+    There are any number of other evaluative metrics for synthetic data.
+We will explore what is available with the Synthetic Data Vault Project that we use and will clarify here as we learn more during experimentation.
+It is worth calling out here that we could expand the evaluation of the synthetic data itself, taking up more scope from later steps such as regression or classification or such.
+The risk of not spending enough time evaluating the synthetic data is that we may end up testing out further modeling on an ill-fitting synthetic dataset.
+
+    Moving on from synthetic data, until we explore the IPEDS data further, I am not sure exactly what we will end up modeling or testing as output.
+But there are some general contenders that we could likely look at.
+
+    For supervised modeling outputs, we can look at standard metrics such as accuracy, precision, F1 score, and more for classification problems.
+We can use a metric like mean squared error (MSE) or adjusted r-squared ($R^2_a$) to measure regression models.
+These evaluations give us firm numbers we can use to determine how well our modeling is doing.
+We may stick with supervised learning if we give more space to synthetic data evaluation since my experience is that the degree of difficulty jumps up when we pivot to unsupervised learning
+
+    That said, I am actually more intrigued by seeing what we can find with unsupervised learning.
+Admittedly, the outputs for unsupervised learning can be more difficult to work with, especially at this early stage.
+Clustering is the first unsupervised modeling that comes to mind.
+I often think of clustering as a much more advanced version of EDA.
+We are not really sure what patterns exist in the data, so we supercharge standard EDA and see what patterns we uncover when leveraging machine learning.
+But we do not have a clean evaluation metric with clustering.
+We will need to interactively explore the clustering outputs, at least at first.
+At some point, we can determine a set number of clusters and pivot the clustering into a classification approach, but the more interesting piece here for me is the clustering itself.
+
+    I am on the fence about leaning more into the synthetic data approach in which we would spend a lot more time evaluating the quality of the synthetic data.
+With synthetic data, I still would like to test out one of the analytical pieces at the end of the pipeline, but maybe we stick to something simpler like a clearer classification problem.
+If we instead drop synthetic data, then we have more room to tackle something like unsupervised learning and the complexity inherent in it.
+
+    While I feel like there are loads of evaluative methods to explore, I am also interested in looking at some of the algorithms from the second data mining course.
+I have leveraged things like similarity scores in other projects, but I have not experimented with, for instance, FP growth, and it would be interesting to possibly pick a brand new algorithm to plop at the end of the pipeline and figure out what it means to evaluate that algorithm and what it reveals to us.
+But, similar to with unsupervised learning, my desire to explore synthetic data may mean leaving out some of the new algorithms from the second course for now.
+This would be an area to explore more in the next iteration of the project though.
 
 \section{Discussion}
-Project timeline
-    - When to finish
-    - Current status
-    - Determine if you need to do less or can do more in the given time
-Potential challenges
-Alternative approaches and backup plan
+    My goal for this project is to stick within a 3-4 week limit, if possible.
+I tend to drag out projects, taking the opportunity to add increased complexity and explore new items I am curious about.
+But, while spending forever on projects is fun and educational, I do risk getting too comfortable with taking my time.
+I could benefit from descoping and focusing on getting smaller chunks working earlier and then deciding if it is worth expanding out from there.
 
-\subsection{Questions for the Reader}
-    \begin{itemize}
-        \item Is adding in the synthetic data component too much?
-I've been exploring this is personal and school projects as well as using it at work, so it's fresh on my mind to keep testing out, but it may be overkill for this project.
-    \end{itemize}
+    In that vein, I am hoping that this draft is an upper bound for the work in this project.
+Some areas where I would like to bound the scope a bit would be in the back-and-forth with the database earlier in the pipeline, how much to dig into synthetic data, how wide of a range of questions to explore in the data, and what types of modeling or pattern mining to use at the end of the pipeline.
+Since I have not started in on the technical work or deeper dives into the data yet, it feels okay to start out more expansively as is done here and look for the next draft to hone in on a more accurate scope.
+
+    I did spend a bit of time looking through different public datasets, a number of those through NCES, ultimately landing on IPEDS, but I still need to spend a bit more time with the chosen IPEDS data.
+I have a sense the three domains of IPEDS data may get trimmed down to two domains.
+But gaining clarity on the data itself will take some time.
+
+    The main challenge I see is that the data ends up being much more complicated to work with than expected, and that will take time to reveal itself, particularly in the understanding and preprocessing steps of the pipeline.
+One goal will be to get to those steps sooner to figure out what it will be like working with this data.
+The next big challenge I foresee is when we hit modeling, both the generative modeling for the synthetic data and whichever modeling we choose to employ on that synthetic data, though, as said above, the goal is for the latter to remain simpler in this project, leaving space to focus on the synthetic data itself.
+
+    The general backup plan is to simplify the IPEDS dataset down quite a bit.
+We may even be able to get away with using just one domain of IPEDS data if needed.
+Since a major goal of this project is to demonstrate synthesizing IPEDS data, we may also only keep data that looks relatively clean to start with.
+We will also get a sense of which features in the data may be easier to model, again providing a way to simplify the scope to accommodate competing needs.
+
+    To wrap up with an explicit acknowledgment of risks for my projects, the biggets worry is that the scope grows to an unruly size, so there are a number of comments in this report speaking to ways to keep that in check, spanning from stricter time scoping to simplifying the data to sticking with easier-to-implement modeling.
 
 \section{Conclusion}
-Conclusion
-Project summary
-Key findings (when finished)
-Future work (when finished)
+    There is not much to conclude at the moment.
+We will fill this in as we start doing the data work of the project.
+What I expect to see here is that we trimmed down the IPEDS data, scoped the use case for the project to be a bit narrower, spent more time explore synthetic data and its associated generative modeling, and picked a simpler overall output, likely sticking supervised modeling.
+For a surprise finish, we may end up deciding to pull some data out and run it through a separate business intelligence tool like Tableau with the idea that end users of the production pipeline may want to leverage tools such as that for their purposes, though this would only come into play if we find that the project turns out to be too quick and simple to get through.
+
+    As for future iterations of the project, I expect that the biggest area of grows would be to incorporate more IPEDS data to answer a broader range of questions.
+I do not expect radical changes in the pipeline architecture in next iterations of the project, but I do expect additions and expansions in order to accommodate particular questions of interest inherent in the new data.
 
@@
 \section{Technical Details}
-
-    Need some details plz
+    Leaving this as a placeholder to pull the technical details out of the narrative above but still keep it available for readers.
 

### 2025-02-18 src/load_data.py
add code to download 2023 versions of files for the three selected domains of ipeds data

@@ -1,0 +1,65 @@
+import requests
+import zipfile
+import io
+
+# Sample URL
+# 'https://nces.ed.gov/ipeds/datacenter/data/IC2023.zip'
+# Institutional characteristics: IC
+# Enrollment: E12
+# Completions: C
+
+ipeds_data_base_url = 'https://nces.ed.gov/ipeds/datacenter/data'
+
+def download_files(files_list, path_files):
+    for f in files_list:
+        # Data files
+        r = requests.get(f'{ipeds_data_base_url}/{f}.zip')
+        if r.ok:
+            with zipfile.ZipFile(io.BytesIO(r.content), 'r') as zipz:
+                zipz.extractall(path_files)
+        else:
+            print(f'bad file: {f}')
+
+        # Dictionaries
+        r = requests.get(f'{ipeds_data_base_url}/{f}_DICT.zip')
+        if r.ok:
+            with zipfile.ZipFile(io.BytesIO(r.content), 'r') as zipz:
+                zipz.extractall(path_files)
+        else:
+            print(f'bad file: {f}_DICT')
+
+
+def load_data():
+    files_institutional_characteristics = [
+        'HD2023'
+        , 'IC2023'
+        , 'IC2023_AY'
+        , 'IC2023_PY'
+        , 'IC2023_CAMPUSES'
+    ]
+    path_files_institutional_characteristics = '../data/institutional_characteristics'
+    download_files(files_institutional_characteristics, path_files_institutional_characteristics)
+
+    files_enrollment = [
+        'EFFY2023'
+        , 'EFFY2023_DIST'
+        , 'EFFY2023_HS'
+        , 'EFIA2023'
+    ]
+    path_files_enrollments = '../data/enrollments'
+    download_files(files_enrollment, path_files_enrollments)
+
+    files_completions = [
+        'C2023_A'
+        , 'C2023_B'
+        , 'C2023_C'
+        , 'C2023DEP'
+    ]
+    path_files_completions = '../data/completions'
+    download_files(files_completions, path_files_completions)
+
+    files_all_survey_components = [
+        'FLAGS2023'
+    ]
+    path_files_all_survey_components = '../data/all_survey_components'
+    download_files(files_all_survey_components, path_files_all_survey_components)

### 2025-02-18 src/orchestrator.py

@@ -1,0 +1,4 @@
+from load_data import load_data
+
+if __name__ == '__main__':
+    load_data()

### 2025-03-08 artifacts/report/msds_data_mining_project_report.tex
add temp additions and notes in latex report after a close call with a git mishap

@@
 %\citestyle{acmauthoryear}
+
+\usepackage{hyperref}
 
@@
 IPEDS is a subset of NCES data that focuses on postsecondary or higher education data.
+IPEDS data is large, complex, and valuable, so it also lends itself well to data mining.
 \cite{ipeds}
@@
 \section{Proposed Work}
-
     The general approach for the work in this project is to leverage Python scripts to move information around, a relational database running in Docker for any pieces of the pipeline that would normally use a database, and notebooks as needed for exploratory and visualization purposes.
@@
 The ideas in this section are fine for now, but we need to make these more concrete as we begin digging into the data.
+
+    \subsection{updates for proposed work}
+        Is synthetic data generative modeling enough, or do we need to keep some extra modeling piece? If the latter, can it be more of a demo or POC than real analysis?
+
+        Can reference technical details section in appendix to fill in from here
+
+        Why synthesize the IPEDS data we use instead of student-level data, the area where we will see a real drop off in coming years? How are aggregates in IPEDS different from student-level data?
+
+        handling different null flags or binary encodings even in same files (good use for slide, talking about data issues in general)
+
+        talk about dimensional-ish modeling (good use for slide)
+
+        goal of testing out mini data mart layer as well for whichever subset we want to synthesize
+
+        call out emphasis on db work since we are mimicing a real-world mining pipeline that will likely rely heavily on db tech
+
+        talk about slowly changing dimensions for institutional characteristics between years in particular and modeling decisions related to that
+
+    \subsection{IPEDS Data}
+        IPEDS has a ton of data available, so our first major task is to narrow that down to what we want to explore in this project.
+The approach we take is to pull in a subset of the three domains of IPEDS data we identified earlier: institutional characteristics, enrollments, and completions.
+There are a ton of trade-offs in terms of pulling in more data or trimming down even to where we do.
+
+        My primary worry with IPEDS data is its complexity.
+Two examples are as follows.
+A number of files in the domains we explore have more than 100 features, and there are many more files across domains that we do not explore, so there are potential concerns related to the size of the data.
+Next, we have similar features across files.
+For example, we have breakdowns by race/ethnicity and gender and across different values for the primary attribute in files, and considering ways to combine these or compare across them are more complicated than they may seem at first.
+
+        The more data we include now, the harder the first pass of modeling will be, so we take the strategy of building out a full pipeline with a small subset of data in such a way that it can be expanded later without too much extra effort.
+Further, since our primary goal is generating synthetic data, we are less worried about finding the most impactful IPEDS features in terms of meaning, so we have some flexibility for the dataset that we choose for now.
+
+        \subsubsection{Chosen Domains}
+
+            Institutional characteristics are the foundation of IPEDS data, providing information related to higher education institutions.
+
+            Next, we use 12-month enrollment data, looking at unduplicated enrollment counts for students who start anytime in a 12-month period and including both degree- and nondegree-seeking students.
+IPEDS also have fall enrollment data, but we choose the wider timeframe.
+Since we also include completions instead of graduations, we capture students who take a wider range of programs.
+12-month enrollment will better capture nondegree students in particular.
+
+            Finally, we use completions, looking at students who complete degree or nondegree credentials.
+This differs from graduation rates that measure students who receive degrees and excluding students who attain nondegree credentials, certificates being a common nondegree credential.
+
+            The goal for now is to continue to narrow down data within each of these domains, building towards a working dataset.
+The general idea is to have one row per institution, and we will need to group across rows in some files to make that happen.
+For example, some files have more than one row per institution, so we need to decide if it is appropriate to smoosh all of the rows per insititution together in those files and sum column values or if we need a more complicated approach.
+We figure these steps out doing the first round of EDA in the database.
+
+        \subsubsection{What's Left from IPEDS}
+
+            IPEDS includes other domains of data, including:
+
+            \begin{itemize}
+                \item Cost
+                \item Financial aid
+                \item Finance
+                \item Admissions
+                \item Outcome Measures
+                \item Graduation rates: students who graduate with a degree within 150\% of the estimated time for program completion
+                \item Graduation rates 200: students who graduate with a degree within 200\% of the estimated time for program completion
+                \item Fall enrollment (as discussed above)
+            \end{itemize}
+
+            Each of these domains has multiple data files, similar to the three domains we are working with.
+The numbers of questions we can ask of IPEDS data multiplies when we pull in these other domains, but we will save those questions for future iterations of this project.
+
+            These other domains may also pose questions related to synthesizing data for them.
+
+        \subsubsection{Leftover Thoughts}
+
+            NCES provides preprocessing before making IPEDS data files available, including tasks such as imputation for missing values, creation of new features for estimated values to compare to reported values, and more.
+We will not worry about these initial preprocessing steps for this iteration of the project.
+
+            For each of the files we keep, we remove gender and race/ethnicity disaggregates.
+These are important features for answering meaning questions with IPEDS, but we remove them in order to simplify the analysis scope in this iteration of the project.
+
+            For a more detailed summary of the IPEDS domains we work with, please read data/data\_notes.md in the GitHub repository.
 
@@
 
+    \subsubsection{Warehousing}
+    \subsubsection{Generative Modeling}
+    \subsubsection{Data Analysis or Business Intelligence}
+
 \section{Conclusion}
@@
 \section{Technical Details}
-    Leaving this as a placeholder to pull the technical details out of the narrative above but still keep it available for readers.
+
+    figure out what should go here vs in proposed work section
+
+    This section will provide more details about the technical work, if of interest. You can find the source code for this project at \href{https://github.com/jhleakakos/msds-data-mining-project}.
+
+    src/orchestrator.py is the primary Python file, controlling the general flow of the pipeline.
+We do not need a full orchestration solution for this project, so we will recreate the idea, as needed, in this file, not worrying too much about automation but instead using this utility to help iterative work.
+One area of growth for larger or more complex systems would be to build a more complex orchestration file and infrastructure or incorporate orchestration software instead.
+
+    The first milestone is getting the subset of IPEDS data that we want to work with loaded into the database.
+Since many real-world pipelines will start with data in a database or loading raw data into a database, we will recreate that here.
+For now, we run the database scripts manually.
+
+    \begin{itemize}
+        \item Download IPEDS flat files from the IPEDS website using src/pull_data.py.
+This script has code to pull all the files in the three domains of IPEDS data that we explore in this project.
+I originally downloaded and looked through all these files and their associated dictionaries, then reducing down to the files and features we want to explore.
+We comment out files that we do not plan to use in this iteration of the project.
+This script is easy to modify to pull in other IPEDS data flat files.
+        \item Set up PostgreSQL to run in Docker.
+The root of the repository includes a Dockerfile that primarily copies the IPEDS files into the container along with a Docker Compose file that configures the remainder of the container and data volume.
+        \item There are structured query language (SQL) scripts to create the datalake database, create the data lake tables, and copy in full IPEDS files as a stepping stone to pulling out the features in each file that we want to work keep.
+        \item We load the data lake by using the \textit{copy} command.
+This takes in a comma-separated values (CSV) file and loads it into the specified table.
+This is purposely different than a process like reading the data into Python structures, opening a database connection, and using Python to write to the database.
+The idea here is that this method will let us load arbitrarily large datasets through a more efficient database utility.
+        \item For now, we run the SQL scripts manually, though we could look to automate these later, likely through storing the data lake loading functionality as a stored procedure or setting up triggers to copy data around.
+    \end{itemize}
+
+    At this stage, we are ready to pivot to EDA and take a closer look at the subset of IPEDS data that we hope to eventually synthesize and feed into modeling.
+
+\section{IPEDS Fields}
+
+    The summaries for files and fields here come from each file's associated data dictionary.
+Note that the biggest difference between the list here and the larger summary in data/data_notes.md in the source repository is that we drop gender and race/ethnicity disaggregates here.
+
+    Domain: Institutional Characteristics
+
+    HDyyyy: directory information for institutions
+
+    \begin{itemize}
+        \item UNITID: unique institutional identifier
+        \item INSTNM: institution name
+        \item STABBR: state abbreviation
+        \item OBEREG: Bureau of Economic Analysis regions, grouped by homogeneity of states based on economic, demographic, social, and cultural characteristics
+        \item HLOFFER: highest level of credential offered
+        \item DEGGRANT: does the institution offer degrees (as compared to non-degree credentials)
+        \item HBCU: is the institution an HBCU
+        \item TRIBAL: is this a tribal institution
+        \item LOCALE: geographic status of school
+        \item CLOSEDAT: date institution closed
+        \item CYACTIVE: is institution active in current year
+        \item INSTCAT: category of institution regarding credentials offered
+        \item C21BASIC: alternative to Carnegie Classification and last updated in 2021
+        \item INSTSIZE: enrollment size in 2019
+        \item CBSA: core based statistical area
+    \end{itemize}
+
+    ICyyyy: program and award level offerings, control, and affiliation of institutions
+
+    \begin{itemize}
+        \item UNITID: unique institutional identifier
+        \item CNTLAFFI: institution control or affiliation (public, private non-profit, etc)
+        \item OPENADMP: open admission policy
+    \end{itemize}
+
+    ICyyyy_PY: student charges by program
+
+    \begin{itemize}
+        \item UNITID: unique institutional identifier
+        \item PRGMOFR: number of programs offered at institution
+        \item CIPCODE1: CIP code of the largest program
+        \item CIPCODE2: CIP code of the second largest program
+        \item CIPCODE3: CIP code of the third largest program
+        \item CIPCODE4: CIP code of the fourth largest program
+        \item CIPCODE5: CIP code of the fifth largest program
+        \item CIPCODE6: CIP code of the sixth largest program
+    \end{itemize}
+
+    Domain: Enrollments
+
+    EFFYyyyy: unduplicated 12-month enrollment headcount
+
+    \begin{itemize}
+        \item UNITID: unique institutional identifier
+        \item EFFYALEV: level, full- and part-time status, degree- or nondegree-seeking status, and year of study of student enrolled for credit
+        \item EFFYLEV: level of study
+        \item EFYTOTLT: total students enrolled for credit
+    \end{itemize}
+
+    EFFYyyyy_DIST: unduplicated 12-month enrollment headcount by distance education status and level of student
+
+    \begin{itemize}
+        \item UNITID: unique institutional identifier
+        \item EFFYDLEV: level and degree- or nondegree-seeking status of students enrolled for credit
+        \item EFYDETOT: all students enrolled for credit (sum of next three columns)
+        \item EFYDEEXC: students enrolled only in distance education courses
+        \item EFYDESOM: students enrolled in at least one distance education course but not enrolled exclusively in distance education courses
+        \item EFYDENON: students not enrolled in any distance education courses
+    \end{itemize}
+
+    EFFYyyyy_HS: unduplicated 12-month enrollment headcount of dual credit students; subset of the nondegree and non-certificate-seeking students from EFFY2023
+
+    \begin{itemize}
+        \item UNITID: unique institutional identifier
+        \item EFYTOTLT: total dual credit students
+    \end{itemize}
+
+    EFIAyyyy: instructional activity measured in total credit and/or contact hours delivered by institutions in 12-month period, used to derive 12-month full time equivalent (FTE) enrollments
+
+    \begin{itemize}
+        \item UNITID: unique institutional identifier
+        \item EFTEUG: estimated FTE undergraduate enrollment
+        \item EFTEGD: estimated FTE graduate enrollment
+        \item FTEUG: reported FTE undergraduate enrollment
+        \item FTEGD: reported FTE graduate enrollment
+        \item FTEDPP: doctor's-professional practice FTE
+    \end{itemize}
+
+    Domain: Completions
+
+    Cyyyy_A: number of awards by type of program, level of award (degree or certificate), first or second major, by race/ethnicity, and by gender
+
+    \begin{itemize}
+        \item UNITID: unique institutional identifier
+        \item CIPCODE: CIP code
+        \item MAJORNUM: first or second major
+        \item AWLEVEL: award level code
+        \item CTOTALT: awards/degrees conferred to all recipients
+    \end{itemize}
+
+    Cyyyy_B: number of students who completed any degree or certificate by race/ethnicity and gender
+
+    \begin{itemize}
+        \item UNITID: unique institutional identifier
+        \item CSTOTLT: number of students receiving awards/degrees
+    \end{itemize}
+
+    Cyyyy_C: number of students receiving a degree or certificate by level of award and race/ethnicity, gender, and age
+
+    \begin{itemize}
+        \item UNITID: unique institutional identifier
+        \item AWLEVELC: award level
+        \item CSTOTLT: number of students receiving awards/degrees
+    \end{itemize}
+
+    CyyyyDEP: number of programs offered by type of program, award level, and distance education status
+
+    \begin{itemize}
+        \item UNITID: unique institutional identifier
+        \item CIPCODE: six-digit CIP code
+        \item PTOTAL: number of programs offered across all award levels
+        \item PTOTALDE: number of programs that can be completed entirely online across all award levels
+        \item PASSOC: number of programs offered at associate's degree level
+        \item PBACHL: number of programs offered at bachelor's degree level
+        \item PMASTR: number of programs offered at master's degree level
+        \item PDOCRS: number of programs offered at doctor's research/scholarship degree level
+        \item PDOCPP: number of programs offered at doctor's professional practice degree level
+        \item PDOCOT: number of programs offered at doctor's other degree level
+        \item PCERT1A: number of programs offered at certificate level of less than 300 clock hours, 9 semester/trimester credit hours, or 13 quarter credit hours
+        \item PCERT1B: number of programs offered at certificate level of 300-800 clock hours, 9-29 semester/trimester credit hours, or 13-44 quarter credit hours
+        \item PCERT2: number of programs offered at certificate level of one year but less than two years
+        \item PCERT4: number of programs offered at certificate level of two years but less than four years
+        \item PPBACC: number of programs offered at postbaccalaureate certificate level
+        \item PPMAST: number of programs offered at post-master's certificate level
+    \end{itemize}
 

### 2025-03-08 src/orchestrator.py
add script to create and populate warehouse tables

@@ -2,3 +2,6 @@ from pull_data import pull_data
 
 if __name__ == '__main__':
     pull_data()
+
+    # I am running the database scripts manually, but we could trigger them from here, probably with a
+    # service account that can call stored procedures that encapsulate the scripts in src/db

### 2025-03-09 artifacts/report/msds_data_mining_project_report.tex
add file summaries to latex report

@@
 
+    \subsection{additions for conclusion}
+
+    This approach is probably over-engineered for the scope of this project, but it lays the groundwork for pulling in much more IPEDS data and running more complicated analyses
+
 \bibliographystyle{ACM-Reference-Format}
@@
 \section{IPEDS Fields}
-
     The summaries for files and fields here come from each file's associated data dictionary.
@@
 
+    \subsection{IPEDS Fields for Modeling}
+        These are the fields we pull from the files above to create the modeling dataset.
+
+        \begin{itemize}
+            \item year: we snag this when reading in files and add it as needed throughout the warehouse
+            \item HDyyyy.UNITID: unique institutional identifier
+            \item HDyyyy.INSTNM: institution name
+            \item HDyyyy.STABBR: state abbreviation
+            \item HDyyyy.OBEREG: Bureau of Economic Analysis regions, grouped by homogeneity of states based on economic, demographic, social, and cultural characteristics
+            \item HDyyyy.HLOFFER: highest level of credential offered
+            \item HDyyyy.DEGGRANT: does the institution offer degrees (as compared to non-degree credentials)
+            \item HDyyyy.HBCU: is the institution an HBCU
+            \item HDyyyy.TRIBAL: is this a tribal institution
+            \item HDyyyy.LOCALE: geographic status of school
+            \item HDyyyy.CLOSEDAT: date institution closed
+            \item HDyyyy.INSTCAT: category of institution regarding credentials offered
+            \item HDyyyy.CBSA: core based statistical area
+            \item ICyyyy.CNTLAFFI: institution control or affiliation (public, private non-profit, etc)
+            \item EFFYyyyy.EFYTOTLT: total students enrolled for credit
+            \item Cyyyy_B.CSTOTLT: number of students receiving awards/degrees
+        \end{itemize}
+
 \end{document}

### 2025-03-14 artifacts/report/msds_data_mining_project_report.tex
rework start of proposed work section

@@
 
-    But, this project will instead focus on a different aspect of declining higher education enrollments: reduced data.
+    This project will instead focus on a different aspect of declining higher education enrollments: reduced data.
 As student enrollments drop and HEIs close or shrink, we are likely to see reduced amounts of higher education data available for analysis, prediction, evaluation, and more.
@@
 
-    Synthetic data is ``artificially generated [data] that resemble[s] the actual data -- more precisely, having similar statistical properties'' ~\cite[p.
+    Synthetic data is ``artificially generated [data] that resemble[s] the actual data --- more precisely, having similar statistical properties'' ~\cite[p.
 1]{decristofaro2024syntheticdatamethodsuse}.
@@
 
-    Building and training models to generate synthetic tabular data -- tabular meaning data in table form with rows and columns -- is notoriously difficult.
+    Building and training models to generate synthetic tabular data --- tabular meaning data in table form with rows and columns --- is notoriously difficult.
 We will leverage existing models and research for the synthetic generation aspect of the pipeline in particular.
@@
 If we can determine trends now, we may be able to utilize synthetic data in the future with dwindling real data to confirm continued trends or to identify changes in those trends.
+Since the focus here is on synthetic data, we will spend more of our time testing out data mining techniques on the real and synthetic data and confirming that we see similar results, supporting use of synthetic data as a stand-in for or supplement to real data.
 
-    To summarize, this project will work with IPEDS data with its focus on higher education and will explore how we can synthesize more IPEDS data to address potential future lack-of-data issues due to declining higher education enrollments, and we will use these data to tackle standard data mining questions related but focused on higher education data.
+    To summarize, this project will work with IPEDS data with its focus on higher education and will explore how we can synthesize more IPEDS data to address potential future lack-of-data issues due to declining higher education enrollments, and we will use these data to tackle standard data mining questions focused on higher education data but across real and synthetic data.
 
@@
 Each KDE can model one mode of the data.
-We then use Gaussian mixture methods (GMMs) to sample from more than one distribution -- each distribution coming from one of those built with the KDEs -- to recreate the actual distribution.
+We then use Gaussian mixture methods (GMMs) to sample from more than one distribution --- each distribution coming from one of those built with the KDEs --- to recreate the actual distribution.
 We then sample from this GMM when we are generating a value for that feature, and this gets baked into the modeling, training, and generation processes.
@@
 Note that both TGAN and CTGAN are dedicated repositories in the Synthetic Data Vault project.
-There is another main library -- sdv -- that provides easier-to-use versions of the models.
+There is another main library --- SDV --- that provides easier-to-use versions of the models.
 There is also SDGym, a synthetic data evaluation library that we can lean on to determine how well our generative modeling is doing.
@@
 So, the contribution that we provide is to start a discussion around data mining for IPEDS data along with generative modeling for IPEDS data.
-We build on the ideas in the papers above -- the general approaches to analysis in the earlier papers, and the use of synthetic tabular data as described in the latter papers.
+We build on the ideas in the papers above --- the general approaches to analysis in the earlier papers, and the use of synthetic tabular data as described in the latter papers.
 
 \section{Proposed Work}
-    The general approach for the work in this project is to leverage Python scripts to move information around, a relational database running in Docker for any pieces of the pipeline that would normally use a database, and notebooks as needed for exploratory and visualization purposes.
 
-    To snag the IPEDS data, we will use Python scripts to download the data from the IPEDS website and store it in a relational database, possibly storing the data on disk as a stepping stone.
-The database will run locally in Docker.
-IPEDS data already comes in a relatively relational form, so there may not be too much work for the initial load into the database.
-Much of the data we end up working with in the real world lives in a database, so this recreates that reality here.
+    The general approach for the work in this project is to recreate a full data pipeline on a smaller scale to demonstrate what we could build out further in a production system.
+We use Python scripts to simulate pulling data from application programming interfaces (APIs), run databases in Docker --- including a very small scale warehouse --- to stand in for real-world database usage, and leverage Jupyter Notebooks for exploratory data analysis (EDA) and modeling, including our core focus of generative modeling to create synthetic tabular data.
+We hold off on saving the final tuned versions of models and putting them into automated pipelines, but that would be an extension that we would include in a production system.
+The main idea is to have all the pieces that we would normally tap into in a regular-enough end-to-end data system to provide a foundation from which to expand on later.
 
-    We may leverage the database for preprocessing.
-We would need to do some manual exploration first in Python, but, after that, we could look at shifting preprocessing into structured query language (SQL), particularly stored as insert triggers in the database.
-Right now, there is a question about how that would intersect with exploratory data analysis (EDA) that we would end up doing in Python, but we can figure out the specifics as we dig into the data.
-My thought here is that a production pipeline would be simpler if we could handle more of the preprocessing automatically in the database and not require that someone hook into separate Python scripts, whether automated or needing to be run manually.
+    \subsection{Pipeline Walkthrough}
 
-    We will need to go through a full EDA process.
-This step will happen outside of the database, so we will look at Python scripts here.
-At first, this is likely easier with the read-eval-print loop (REPL) functionality of a notebook, though we do not want notebooks in our final pipeline until much closer to the end in production.
-We need the automation power in scripts instead of notebooks.
-We may also learn more from the EDA that we can bake into extra preprocessing, informing what is mentioned in the last paragraph.
+    You can view the source code at \href{https://github.com/jhleakakos/msds-data-mining-project}.
 
-    After this round of EDA and any further preprocessing we glean from it, we can then think about warehousing.
-We will shift the data back into the database, likely just matching the new form of the data to what we need for modeling, but we can decide later on regarding the form of the data in the warehouse.
-For the scope of this project, we will likely end up with a bespoke warehouse built to immediate modeling or analysis needs rather than a broader warehouse set up in something like a full star schema.
+    We start by using a Python script as a placeholder for some sort of orchestration system.
+Separate orchestration software would be overkill right now, so we just import functions related to pulling data from the IPEDS website and call those from a dedicated orchestration script, but we would at some point possibly look into something like Apache Airflow to handle orchestration.
+
+    IPEDS data already comes in a relatively relational format, and that lends itself well to loading it directly into a relational database serving as a data lake.
+We often work with databases in real-world projects, so it makes sense to include one here.
+We run a PostgreSQL database locally in Docker.
+Because Postgres connections are database-specific, we use one shared databse for this layer of the data lake and the coming data warehouse layer, using schemas to separate out the data lake from the data warehouse.
+This allows us to use structured query language (SQL) scripts to move data between the data lake and data warehouse without needing to use any other tooling, simplifying the handling of data in the database.
+
+    We pull a subset of data out of the data lake and explore it in a Jupyter Notebook.
+The main goal here is to understand the already trimmed IPEDS data and determine where we want to start in terms of what gets moved into the warehouse.
+IPEDS data is vast and often difficult to work with, so it is important at this stage to narrow down the scope of the data to something reasonable and that allows us to keep our focus on the coming modeling.
+One big area of focus is on looking at null or missing flags in the IPEDS data and deciding how to handle those when we move data into the warehouse.
+
+    We then use SQL script to move data from the data lake into a very rudimentary version of a dimensional model in the data warehouse, leveraging a standard star schema.
+Again, this is super small compared to what you would normally see in a real warehouse, but it lets us start to explore different ways of working with IPEDS data on a small scale before ramping up size and complexity in future itertations.
+As stated above, we use schemas in a shared database to separate out the data lake and data warehouse.
+
+    Note that we do all the database work manually in SQL scripts.
+This is one of the first areas for improvement in future iterations.
+An easy start would be to store the current SQL scripts as stored procedures and call them either from an orchestration script or from update or insert triggers in the database.
+I went back and forth on adding some of these in, but, for the focus of the current project, it felt more important to keep an eye on the modeling instead of adding further complexity into the database layers.
+But this is probably the next improvement I would make, along with the improvement at the end of modeling mentioned later.
+
+    Next we pull data out of the warehouse for modeling.
+We do another pass of EDA to explore how quirks in the data may affect generative modeling.
+We do see some issues and will speak about those in evaluation later.
+
+    We then start in on the main focus of the project --- generative modeling to create new tabular synthetic data that can stand in for or supplement IPEDS data.
+As stated earlier, this type of generative modeling is notoriously difficult, so, since the focus of this project is not on building an optimized generative model but instead on generating new IPEDS data, we leverage already existing libraries for the generative modeling --- the Synthetic Data Vault project and its CTGAN generative model in particular.
+I had decided early on to use CTGAN for generative modeling since I spent some time exploring the associated CTGAN paper in a previous project, but there are a number of models and ways to experiment with them, and I tested each of those out before pivoting to working with CTGAN.
+
+    While the notebook for generative modeling is not too long in its current state, much of my time outside of understanding the IPEDS data has gone into experimenting with the generative modeling.
+The SDV library's APIs are relatively straightforward to work with, but there are many hyperparameters to play around with.
+There are also decisions about dropping some data features to facilitate quicker and more accurate generative modeling, dropping categorical features with too many categories in particular.
+
+    I used feedback from generative modeling steps and from my understanding of what could negatively affect generative modeling to help further trim down the feature space of the IPEDS data used for modeling.
+For example, we cut out some of the categorical variables that would be interesting to explore in terms of meaning --- race/ethnicity and gender as two major groups of categorical variables --- because these increase the feature space and copmlexity of inter-feature relationships in a way that can abstract from the needs of the first pass of this project --- feeling like we have landed on a decision about if using our approach to synthetic IPEDS data generation is feasible.
+
+    While working on generative modeling, we also use the SDMetrics library to evaluate the generative modeling process and outputs.
+For example, we are able to provide a summary number for the quality of the synthetic data in terms of how close it is to the real data.
+We are able to visualize the GAN losses when training the CTGAN model.
+We are also able to visualize a given feature in both the real and synthetic data or pairs of features to see different angles into how similar or dissimilar the synthetic data is to or from the real data.
+
+    Is synthetic data generative modeling enough, or do we need to keep some extra modeling piece? If the latter, can it be more of a demo or POC than real analysis?
 
@@
     \subsection{updates for proposed work}
-        Is synthetic data generative modeling enough, or do we need to keep some extra modeling piece? If the latter, can it be more of a demo or POC than real analysis?
 
@@
         talk about slowly changing dimensions for institutional characteristics between years in particular and modeling decisions related to that
+
+        first eda from data lake to warehouse shows that the expected groupings aren't how the appear.
+for example, when you sum for all the enrollment rows in effy2023.csv for an institution, the total is way higher than is reasonable and than what we see on the IPEDS data explorer tool on their website.
+The dictionaries aren't super clear about how this should happen or how to properly use all of the rows.
+
+        use IPEDS data explorer tool to sanity check the numbers we use for modeling.
+Our numbers are higher than what we see in that tool, but we are using 12-month enrollment compared to the tool's fall enrollment, so we are likely close enough.
+
+        our enrollment numbers are higher, but our top 10 by enrollment match what we see on ipeds and other lists of schools by enrollment
+
+        talk about using single-table generative models from SDV, simplifying first pass of modeling
+
+        talk about not wanting synthetic data to be 100 percent accurate due to desired trade-offs for privacy
+
+        increases from high 70 percent to above for scores when dropping institution name and CBSA due to high number of categories
+
+        Talk about: data understanding (including stats and vis), data preprocessing and steps taken on raw data, emphasize that we do use warehousing and how
+
+        Talk about modeling, probably adding a new subsection to go into detail on synthetic data and detail on regression/classification/clustering
+
+        Pattern evaluation: effectiveness, efficiency
+
+        Emphasize where the project currently is at when updating report and slides
+
+        Emphasize analytical thinking, reasoning behind choices, process, method, results
+
+        Talk about how to evaluate generative output -- machine learning score and how that fits in with other outputs
+
+        pick some subsections for this section to clarify items here
 
@@
 \section{Evaluation}
+
+    talk about issues in the data and curiosity around if generative modeling will handle them okay -- skewed data and imbalanced categories
+
+    SDMetrics
+
+    Summary model score
+    SDMetrics numbers or visualizations
+
     The first evaluations will capture how similar the synthetic data is to the real data.
@@
 \section{Discussion}
+
     My goal for this project is to stick within a 3-4 week limit, if possible.
@@
 \section{Conclusion}
+
+    This approach is probably over-engineered for the scope of this project, but it lays the groundwork for pulling in much more IPEDS data and running more complicated analyses
+
     There is not much to conclude at the moment.
@@
 
-    \subsection{additions for conclusion}
-
-    This approach is probably over-engineered for the scope of this project, but it lays the groundwork for pulling in much more IPEDS data and running more complicated analyses
-
 \bibliographystyle{ACM-Reference-Format}
@@
 \appendix
-
-\section{Technical Details}
-
-    figure out what should go here vs in proposed work section
-
-    This section will provide more details about the technical work, if of interest. You can find the source code for this project at \href{https://github.com/jhleakakos/msds-data-mining-project}.
-
-    src/orchestrator.py is the primary Python file, controlling the general flow of the pipeline.
-We do not need a full orchestration solution for this project, so we will recreate the idea, as needed, in this file, not worrying too much about automation but instead using this utility to help iterative work.
-One area of growth for larger or more complex systems would be to build a more complex orchestration file and infrastructure or incorporate orchestration software instead.
-
-    The first milestone is getting the subset of IPEDS data that we want to work with loaded into the database.
-Since many real-world pipelines will start with data in a database or loading raw data into a database, we will recreate that here.
-For now, we run the database scripts manually.
-
-    \begin{itemize}
-        \item Download IPEDS flat files from the IPEDS website using src/pull_data.py.
-This script has code to pull all the files in the three domains of IPEDS data that we explore in this project.
-I originally downloaded and looked through all these files and their associated dictionaries, then reducing down to the files and features we want to explore.
-We comment out files that we do not plan to use in this iteration of the project.
-This script is easy to modify to pull in other IPEDS data flat files.
-        \item Set up PostgreSQL to run in Docker.
-The root of the repository includes a Dockerfile that primarily copies the IPEDS files into the container along with a Docker Compose file that configures the remainder of the container and data volume.
-        \item There are structured query language (SQL) scripts to create the datalake database, create the data lake tables, and copy in full IPEDS files as a stepping stone to pulling out the features in each file that we want to work keep.
-        \item We load the data lake by using the \textit{copy} command.
-This takes in a comma-separated values (CSV) file and loads it into the specified table.
-This is purposely different than a process like reading the data into Python structures, opening a database connection, and using Python to write to the database.
-The idea here is that this method will let us load arbitrarily large datasets through a more efficient database utility.
-        \item For now, we run the SQL scripts manually, though we could look to automate these later, likely through storing the data lake loading functionality as a stored procedure or setting up triggers to copy data around.
-    \end{itemize}
-
-    At this stage, we are ready to pivot to EDA and take a closer look at the subset of IPEDS data that we hope to eventually synthesize and feed into modeling.
 
@@
         These are the fields we pull from the files above to create the modeling dataset.
+Note that we drop INSTNM and CBSA during generative modeling and later since these features have a ton of categories each, and all those categories wreak havoc on the CTGAN performance without providing enough meaning to justify the performance hit.
+For context, the CTGAN model takes more than 30 minutes to train with these features and under a minute without them.
 
@@
             \item HDyyyy.UNITID: unique institutional identifier
-            \item HDyyyy.INSTNM: institution name
+            \item HDyyyy.INSTNM: institution name (not using during actual modeling because of the number of unique categories)
             \item HDyyyy.STABBR: state abbreviation
@@
             \item HDyyyy.INSTCAT: category of institution regarding credentials offered
-            \item HDyyyy.CBSA: core based statistical area
+            \item HDyyyy.CBSA: core based statistical area (not using during actual modeling because of the number of unique categories)
             \item ICyyyy.CNTLAFFI: institution control or affiliation (public, private non-profit, etc)

### 2025-03-15 artifacts/report/msds_data_mining_project_report.tex
add ipeds data section in proposed work

@@
     \subsection{Pipeline Walkthrough}
-
-    You can view the source code at \href{https://github.com/jhleakakos/msds-data-mining-project}.
+        You can view the source code at \href{https://github.com/jhleakakos/msds-data-mining-project}{https://github.com/jhleakakos/msds-data-mining-project}.
 
@@
         While working on generative modeling, we also use the SDMetrics library to evaluate the generative modeling process and outputs.
-For example, we are able to provide a summary number for the quality of the synthetic data in terms of how close it is to the real data.
-We are able to visualize the GAN losses when training the CTGAN model.
-We are also able to visualize a given feature in both the real and synthetic data or pairs of features to see different angles into how similar or dissimilar the synthetic data is to or from the real data.
+For example, we are able to provide a summary number for the quality of the synthetic data in terms of how close it is to the real data, we are able to visualize the GAN losses when training the CTGAN model, and we are able to visualize a given feature in both the real and synthetic data or pairs of features to see different angles into how similar or dissimilar the synthetic data is to or from the real data.
 
-    Is synthetic data generative modeling enough, or do we need to keep some extra modeling piece? If the latter, can it be more of a demo or POC than real analysis?
+        Once we have our synthetic dataset, we move on to the final stages of the project.
+I am still working through this part of the project, but the plan is to spend some time testing out different modeling techniques on the real data, the synthetic data, and a combination of both datasets to see if we get similar outputs on all three.
+We have two numeric features that lend themselves to regression.
+We could pick some of the categorical features and experiment with classifications.
+And we could decide to branch out to something more complex such as clustering or association mining.
 
-    The next step is to explore the outputs of the pipeline.
-It occurs to me that there are two competing outputs in this project: synthetic IPEDS data and standard evaluative findings.
-We will need to find a balance here, so we may limit the amount of regression, classification, or clustering we do to leave space for the synthetic data generative modeling, possibly demoing a smaller example of how we would glean insights out of the synthetic dataset using classification, for example, rather than adding more scope into that outputs after the generative modeling.
-The final outputs I have a sense will work well with the synthetic IPEDS data are regression for number of enrollments or degree completions, clustering to find groupings of institutions that we should keep an eye on, and classification for putting institutions into those groupings.
-
-    I have been exploring synthetic data in personal and school projects as well as using it at work, so it is fresh on my mind, and I would really like to include it here, but, as said above, it is in tension with what we might see in a standard data mining pipeline.
-We will need to pull data out of the warehouse, import the synthetic data libraries, run the generative modeling, and then possibly load the synthetic data back into the database.
-This area has a bit of a question mark as to how much scope it will take up.
-I feel comfortable with how the generative modeling will work, but we will need to explore performance with IPEDS data in particular.
-
-    The areas I am keeping an eye on here are the extra loop of preprocessing in the database, the inclusion of synthetic data, and narrowing down the final outputs.
-The ideas in this section are fine for now, but we need to make these more concrete as we begin digging into the data.
-
-    \subsection{updates for proposed work}
-
-        Can reference technical details section in appendix to fill in from here
-
-        Why synthesize the IPEDS data we use instead of student-level data, the area where we will see a real drop off in coming years? How are aggregates in IPEDS different from student-level data?
-
-        handling different null flags or binary encodings even in same files (good use for slide, talking about data issues in general)
-
-        talk about dimensional-ish modeling (good use for slide)
-
-        goal of testing out mini data mart layer as well for whichever subset we want to synthesize
-
-        call out emphasis on db work since we are mimicing a real-world mining pipeline that will likely rely heavily on db tech
-
-        talk about slowly changing dimensions for institutional characteristics between years in particular and modeling decisions related to that
-
-        first eda from data lake to warehouse shows that the expected groupings aren't how the appear.
-for example, when you sum for all the enrollment rows in effy2023.csv for an institution, the total is way higher than is reasonable and than what we see on the IPEDS data explorer tool on their website.
-The dictionaries aren't super clear about how this should happen or how to properly use all of the rows.
-
-        use IPEDS data explorer tool to sanity check the numbers we use for modeling.
-Our numbers are higher than what we see in that tool, but we are using 12-month enrollment compared to the tool's fall enrollment, so we are likely close enough.
-
-        our enrollment numbers are higher, but our top 10 by enrollment match what we see on ipeds and other lists of schools by enrollment
-
-        talk about using single-table generative models from SDV, simplifying first pass of modeling
-
-        talk about not wanting synthetic data to be 100 percent accurate due to desired trade-offs for privacy
-
-        increases from high 70 percent to above for scores when dropping institution name and CBSA due to high number of categories
-
-        Talk about: data understanding (including stats and vis), data preprocessing and steps taken on raw data, emphasize that we do use warehousing and how
-
-        Talk about modeling, probably adding a new subsection to go into detail on synthetic data and detail on regression/classification/clustering
-
-        Pattern evaluation: effectiveness, efficiency
-
-        Emphasize where the project currently is at when updating report and slides
-
-        Emphasize analytical thinking, reasoning behind choices, process, method, results
-
-        Talk about how to evaluate generative output -- machine learning score and how that fits in with other outputs
-
-        pick some subsections for this section to clarify items here
+        There are two competing outputs in this project.
+The focus is on creating synthetic data, but we do not want to lose sight of making sure that the data we synthesize could lead to useful insights.
+I have been exploring synthetic data in personal and school projects as well as using it at work, so it is fresh on my mind, and I would really like to include it here, but it is in tension with what we might see in a standard data mining pipeline.
+We will figure out the balance for the next report update.
 
     \subsection{IPEDS Data}
+
+        Note: you can find a more detailed explanation of the IPEDS data we use in the appendix, and you can find a more detailed summary of the IPEDS domains we work with in data/data\_notes.md in the GitHub repository.
+
         IPEDS has a ton of data available, so our first major task is to narrow that down to what we want to explore in this project.
 The approach we take is to pull in a subset of the three domains of IPEDS data we identified earlier: institutional characteristics, enrollments, and completions.
-There are a ton of trade-offs in terms of pulling in more data or trimming down even to where we do.
+There are many trade-offs in terms of pulling in more data or trimming down even to where we do.
 
@@
 Two examples are as follows.
-A number of files in the domains we explore have more than 100 features, and there are many more files across domains that we do not explore, so there are potential concerns related to the size of the data.
-Next, we have similar features across files.
-For example, we have breakdowns by race/ethnicity and gender and across different values for the primary attribute in files, and considering ways to combine these or compare across them are more complicated than they may seem at first.
+First, a number of files in the domains we explore have more than 100 features, and there are many more files across domains that we do not explore, so there are concerns related to the size of the data.
+Second, we have similar features across files that are more complicated to combine or compare than they may seem at first.
+For example, we remove race/ethnicity and gender disaggregates in order to simplify overall count data, but these are important features for answering research questions.
+
+        NCES provides preprocessing before making IPEDS data files available, including tasks such as imputation for missing values, creation of new features for estimated values to compare to reported values, and more.
+We will not worry about these initial preprocessing steps for this iteration of the project.
 
         The more data we include now, the harder the first pass of modeling will be, so we take the strategy of building out a full pipeline with a small subset of data in such a way that it can be expanded later without too much extra effort.
-Further, since our primary goal is generating synthetic data, we are less worried about finding the most impactful IPEDS features in terms of meaning, so we have some flexibility for the dataset that we choose for now.
+Since our primary goal is generating synthetic data, we are less worried about finding the most impactful IPEDS features in terms of meaning in this iteration, giving us some flexibility for the modeling dataset that we choose for now.
 
-        \subsubsection{Chosen Domains}
+        These are the three IPEDS domains that we work with:
 
-            Institutional characteristics are the foundation of IPEDS data, providing information related to higher education institutions.
+        \begin{itemize}
+            \item Institutional characteristics are the foundation of IPEDS data, providing information related to higher education institutions.
+            \item Completions look at students who complete degree or nondegree credentials.
+This differs from graduation rates that measure students who receive degrees and exclude students who attain nondegree credentials, certificates being a common nondegree credential.
+            \item 12-month enrollment data looks at unduplicated enrollment counts for students who start anytime in a 12-month period and including both degree- and nondegree-seeking students.
+IPEDS also has fall enrollment data, but we choose the wider timeframe.
+Since we include completions, a broader scope than graduations, we capture students who take a wider range of programs.
+12-month enrollment will better capture nondegree students in particular.
+        \end{itemize}
 
-            Next, we use 12-month enrollment data, looking at unduplicated enrollment counts for students who start anytime in a 12-month period and including both degree- and nondegree-seeking students.
-IPEDS also have fall enrollment data, but we choose the wider timeframe.
-Since we also include completions instead of graduations, we capture students who take a wider range of programs.
-12-month enrollment will better capture nondegree students in particular.
-
-            Finally, we use completions, looking at students who complete degree or nondegree credentials.
-This differs from graduation rates that measure students who receive degrees and excluding students who attain nondegree credentials, certificates being a common nondegree credential.
-
-            The goal for now is to continue to narrow down data within each of these domains, building towards a working dataset.
-The general idea is to have one row per institution, and we will need to group across rows in some files to make that happen.
-For example, some files have more than one row per institution, so we need to decide if it is appropriate to smoosh all of the rows per insititution together in those files and sum column values or if we need a more complicated approach.
-We figure these steps out doing the first round of EDA in the database.
-
-        \subsubsection{What's Left from IPEDS}
-
-            IPEDS includes other domains of data, including:
+        IPEDS includes other domains of data that we do not use, including:
 
@@
             \item Graduation rates 200: students who graduate with a degree within 200\% of the estimated time for program completion
-                \item Fall enrollment (as discussed above)
+            \item Fall enrollment: students who enroll in the fall, as compared to the 12-month enrollment we use where students enroll at any point during a 12-month period
         \end{itemize}
 
-            Each of these domains has multiple data files, similar to the three domains we are working with.
+        Each of these domains has multiple data files, similar to the three domains we work with.
 The numbers of questions we can ask of IPEDS data multiplies when we pull in these other domains, but we will save those questions for future iterations of this project.
 
-            These other domains may also pose questions related to synthesizing data for them.
+    \subsubsection{Data Lake and Data Warehouse}
+        We will take a moment to talk about some aspects of the database work. We use a database to mimic what we would likely encounter in real-world mining pipelines.
 
-        \subsubsection{Leftover Thoughts}
+        We use two database layers: a data lake and a data warehouse, though the warehouse is a bit of a stretch in terms of the scope we have for it.
 
-            NCES provides preprocessing before making IPEDS data files available, including tasks such as imputation for missing values, creation of new features for estimated values to compare to reported values, and more.
-We will not worry about these initial preprocessing steps for this iteration of the project.
+        We narrow down the entirety of the data in the three chosen IPEDS domain before loading that subset into the data lake.
+This narrowing focuses on pulling out counts that lend themselves to a first pass of generative modeling and removeing features or groupings of features that have a large number of categories.
+We identify counts across any of the raw files along with disaggregates that appear manageable for the first pass of the project.
+The primary change from what we outline in the appendix compared to what we load into the data lake is that we remove race/ethnicity and gender disaggregates.
 
-            For each of the files we keep, we remove gender and race/ethnicity disaggregates.
-These are important features for answering meaning questions with IPEDS, but we remove them in order to simplify the analysis scope in this iteration of the project.
+        After the first pass of EDA, we need to add lookup tables for the categorical features that we choose to keep in the warehouse.
+The IPEDS dictionaries do not make this task as easy as it should be.
+Some features get dropped between the data lake and data warehouse due to lack of clear mappings for categorical values, even after searching for those on the IPEDS and NCES websites.
+Some of the mappings in the data dictionaries are clear, but others do not provide values for the descriptions that they list, so we do our best to try and figure out the mappings.
 
-            For a more detailed summary of the IPEDS domains we work with, please read data/data\_notes.md in the GitHub repository.
+        One example is for boolean features.
+Some features use 1s and 2s in the original data while other features use 1s and 3s.
+We use the initial pass of EDA to determine which category is more prevalent and use a judgement call to map that to what we believe should be most prevalent according to general understanding of higher education data.
+
+        In other cases, we need to use the IPEDS data explorer tool (\href{https://nces.ed.gov/ipeds/use-the-data}{https://nces.ed.gov/ipeds/use-the-data}) to check what values IPEDS has attributed to given institutions.
+For example, the raw data uses numbers for categorical values.
+When we are unclear if an institution is public, nonprofit private, or for-profit private, we can look that institution up on the IPEDS data explorer, check what value IPEDS has attributed to that institution, and then compare that description to the numeric value we find in the raw data.
+We repeat this for each categorical feature, building up mappings that we can add as lookup tables.
+
+        We then perform the second reduction in data, primarily removing a number of enrollment and completion count categories and totals, simplifying the data further and starting to zoom in on a few numeric features that we can use for regression during modeling.
+We map this data into a dimensional-ish model in the warehouse.
+This is mostly meant as a proof of concept, and it mirrors a setup we could likely run into in real-world applications.
+
+        We limit the numeric aggregates we use in the warehouse, only focusing on those that we want to synthesize during generative modeling.
+We also return to the IPEDS data explorer tool to sanity check the totals we get for enrollments and completions.
+It is not immediately clear the best way to group across the data from each of the tables in the data lake, mostly due to the issue of unclear categories for categorical features mentioned in the previous paragraph.
+So we test out different aggregations and compare the sums to what we see in the IPEDS data explorer too.
+Our counts end up higher than what we see in that tool, but we end up in a reasonable ballpark that lets us finish up the warehouse for this iteration of the project.
+For example, the top 10 largest institutions by enrollment in our warehouse match those that we see in external sources.
+
+        We model the dimension table for institutional characteristics as a slowly changing dimension (SCD) since we expect that some of the values in this table could change for a given institution, but those changes are unlikely to be common or frequent.
+We expect enrollment and completion counts to change between years, but we do not typically expect values describing insitutions to change.
+We will not need to leverage this SCD in this pass of the project since we stick with only working with 2023 data, but we will see some changes when we start adding in more data.
+
+        Note that the appendix has details on the features that we work with at the data lake and data warehouse steps.
+
+    \subsubsection{Generative Modeling}
+
+    Talk about modeling, probably adding a new subsection to go into detail on synthetic data and detail on regression/classification/clustering
+    talk about using single-table generative models from SDV, simplifying first pass of modeling
+
+    \subsubsection{Data Analysis or Business Intelligence}
+
+    Is synthetic data generative modeling enough, or do we need to keep some extra modeling piece? If the latter, can it be more of a demo or POC than real analysis?
 
@@
     SDMetrics numbers or visualizations
+
+    Talk about how to evaluate generative output -- machine learning score and how that fits in with other outputs
+
+    Pattern evaluation: effectiveness, efficiency
+
+    talk about not wanting synthetic data to be 100 percent accurate due to desired trade-offs for privacy
+
+    increases from high 70 percent to above for scores when dropping institution name and CBSA due to high number of categories
 
@@
 
+    timeline
+    challenges
+    alternatives and backup plan
+
     My goal for this project is to stick within a 3-4 week limit, if possible.
@@
 
-    \subsubsection{Warehousing}
-    \subsubsection{Generative Modeling}
-    \subsubsection{Data Analysis or Business Intelligence}
-
 \section{Conclusion}
 
+    project summary
+    key findings
+    future work
+
     This approach is probably over-engineered for the scope of this project, but it lays the groundwork for pulling in much more IPEDS data and running more complicated analyses
+
+    Emphasize where the project currently is at when updating report and slides
+
+    Why synthesize the IPEDS data we use instead of student-level data, the area where we will see a real drop off in coming years? How are aggregates in IPEDS different from student-level data?
 
@@
     The summaries for files and fields here come from each file's associated data dictionary.
-Note that the biggest difference between the list here and the larger summary in data/data_notes.md in the source repository is that we drop gender and race/ethnicity disaggregates here.
+Note that the biggest difference between the list here and the larger summary in data/data\_notes.md in the source repository is that we drop gender and race/ethnicity disaggregates here.
 
@@
 
-    ICyyyy_PY: student charges by program
+    ICyyyy\_PY: student charges by program
 
@@
 
-    EFFYyyyy_DIST: unduplicated 12-month enrollment headcount by distance education status and level of student
+    EFFYyyyy\_DIST: unduplicated 12-month enrollment headcount by distance education status and level of student
 
@@
 
-    EFFYyyyy_HS: unduplicated 12-month enrollment headcount of dual credit students; subset of the nondegree and non-certificate-seeking students from EFFY2023
+    EFFYyyyy\_HS: unduplicated 12-month enrollment headcount of dual credit students; subset of the nondegree and non-certificate-seeking students from EFFY2023
 
@@
 
-    Cyyyy_A: number of awards by type of program, level of award (degree or certificate), first or second major, by race/ethnicity, and by gender
+    Cyyyy\_A: number of awards by type of program, level of award (degree or certificate), first or second major, by race/ethnicity, and by gender
 
@@
 
-    Cyyyy_B: number of students who completed any degree or certificate by race/ethnicity and gender
+    Cyyyy\_B: number of students who completed any degree or certificate by race/ethnicity and gender
 
@@
 
-    Cyyyy_C: number of students receiving a degree or certificate by level of award and race/ethnicity, gender, and age
+    Cyyyy\_C: number of students receiving a degree or certificate by level of award and race/ethnicity, gender, and age
 
@@
             \item EFFYyyyy.EFYTOTLT: total students enrolled for credit
-            \item Cyyyy_B.CSTOTLT: number of students receiving awards/degrees
+            \item Cyyyy\_B.CSTOTLT: number of students receiving awards/degrees
         \end{itemize}

### 2025-03-15 artifacts/report/msds_data_mining_project_report.tex
adjust evaluation section in report

@@
 
-    \subsubsection{Generative Modeling}
+    \subsubsection{Data Analysis}
+        Will fill this section in more for the next pass of the report.
 
-    Talk about modeling, probably adding a new subsection to go into detail on synthetic data and detail on regression/classification/clustering
-    talk about using single-table generative models from SDV, simplifying first pass of modeling
+        This is the remaining chunk of work that I have to complete at this point.
+The current thinking is that we will lean into testing out different modeling or mining techniques across the real data, across the synthetic data, and across both combined together.
+The focus of this project is on synthesizing IPEDS data, so we need to explore how incorporating the synthetic data affects common output tasks that we will want to run on different subsets of IPEDS data.
 
-    \subsubsection{Data Analysis or Business Intelligence}
+        Note that this will end up looking a lot like the machine learning score evaluation technique that we explained earlier.
+This means our modeling analysis will blur the line with evaluation for this project, but it will help us to gain more fluency in working with synthetic data generally and the use of the specific libraries we use with IPEDS data.
 
-    Is synthetic data generative modeling enough, or do we need to keep some extra modeling piece? If the latter, can it be more of a demo or POC than real analysis?
+        As a reminder, we are abstracting many of the details of generative modeling since that could be an entire project to itself, though we will call out how the data we work with could affect the generative modeling step.
 
 \section{Evaluation}
-
-    talk about issues in the data and curiosity around if generative modeling will handle them okay -- skewed data and imbalanced categories
-
-    SDMetrics
-
-    Summary model score
-    SDMetrics numbers or visualizations
-
-    Talk about how to evaluate generative output -- machine learning score and how that fits in with other outputs
-
-    Pattern evaluation: effectiveness, efficiency
-
-    talk about not wanting synthetic data to be 100 percent accurate due to desired trade-offs for privacy
-
-    increases from high 70 percent to above for scores when dropping institution name and CBSA due to high number of categories
-
     The first evaluations will capture how similar the synthetic data is to the real data.
@@
 
-    Moving on from synthetic data, until we explore the IPEDS data further, I am not sure exactly what we will end up modeling or testing as output.
-But there are some general contenders that we could likely look at.
+    At this stage, we have run a pass of generative modeling and added summary statistics and visualizations in the generative modeling notebook that you can explore in the source code.
+We also drop two features that we originally meant to include due to their negative effect on generative modeling.
+I will hold off on doing a more detailed summary until the next pass of this report, though the initial metrics look promising, with the summary numbers falling into ranges that hint that the synthetic data is picking up on patterns in the real data without mirroring the real data so closely that we run into privacy concerns.
 
-    For supervised modeling outputs, we can look at standard metrics such as accuracy, precision, F1 score, and more for classification problems.
-We can use a metric like mean squared error (MSE) or adjusted r-squared ($R^2_a$) to measure regression models.
+    Moving on from synthetic data, my main intention is to build regression models on total enrollment and completion numbers per institution.
+I am curious to see which descriptive features help the most in predicting enrollments and completions.
+This would be a supervised regression task that we have built the warehouse and modeling dataset around.
+We can use metrics like mean squared error (MSE) or adjusted r-squared ($R^2_a$) to measure these regression models.
 These evaluations give us firm numbers we can use to determine how well our modeling is doing.
-We may stick with supervised learning if we give more space to synthetic data evaluation since my experience is that the degree of difficulty jumps up when we pivot to unsupervised learning
+We can compare across the three different groups of input data to see how well the synthetic data is doing, and we can focus on the modeling for the real data to see if we can glean insights out of the IPEDS data.
 
-    That said, I am actually more intrigued by seeing what we can find with unsupervised learning.
-Admittedly, the outputs for unsupervised learning can be more difficult to work with, especially at this early stage.
+    I am not sure yet if we will include unsupervised learning techniques.
+Admittedly, the outputs for unsupervised learning can be more difficult to work with.
 Clustering is the first unsupervised modeling that comes to mind.
@@
 At some point, we can determine a set number of clusters and pivot the clustering into a classification approach, but the more interesting piece here for me is the clustering itself.
-
-    I am on the fence about leaning more into the synthetic data approach in which we would spend a lot more time evaluating the quality of the synthetic data.
-With synthetic data, I still would like to test out one of the analytical pieces at the end of the pipeline, but maybe we stick to something simpler like a clearer classification problem.
-If we instead drop synthetic data, then we have more room to tackle something like unsupervised learning and the complexity inherent in it.
 

### 2025-03-15 artifacts/report/msds_data_mining_project_report.tex
adjust discussion section in report

@@
 \section{Discussion}
-
-    timeline
-    challenges
-    alternatives and backup plan
-
     My goal for this project is to stick within a 3-4 week limit, if possible.
@@
 
-    In that vein, I am hoping that this draft is an upper bound for the work in this project.
-Some areas where I would like to bound the scope a bit would be in the back-and-forth with the database earlier in the pipeline, how much to dig into synthetic data, how wide of a range of questions to explore in the data, and what types of modeling or pattern mining to use at the end of the pipeline.
-Since I have not started in on the technical work or deeper dives into the data yet, it feels okay to start out more expansively as is done here and look for the next draft to hone in on a more accurate scope.
+    The first area to keep an eye on scope is with the data used in the project.
+We start by selecting three IPEDS domains among many.
+We then systematically trim the datasets down in stages, as described earlier.
+By the time we reach modeling, we have a dataset with 14 features.
+Two of those are numeric targets we can use for regression.
+We could also put some boundary cutoffs for those fields to aid with classification.
+We have a couple of more likely features in terms of explanatory power, and we have a few extras to test out in terms of importance.
+Keeping the dataset smaller helps keep the modeling stages more focused.
 
-    I did spend a bit of time looking through different public datasets, a number of those through NCES, ultimately landing on IPEDS, but I still need to spend a bit more time with the chosen IPEDS data.
-I have a sense the three domains of IPEDS data may get trimmed down to two domains.
-But gaining clarity on the data itself will take some time.
+    Next is the use of the database.
+This is a personal thing for me, but I am more familiar working in databases over the years, so doing work in the database simplifies the amount of time needed for me to set up a production-esque system.
 
-    The main challenge I see is that the data ends up being much more complicated to work with than expected, and that will take time to reveal itself, particularly in the understanding and preprocessing steps of the pipeline.
-One goal will be to get to those steps sooner to figure out what it will be like working with this data.
-The next big challenge I foresee is when we hit modeling, both the generative modeling for the synthetic data and whichever modeling we choose to employ on that synthetic data, though, as said above, the goal is for the latter to remain simpler in this project, leaving space to focus on the synthetic data itself.
+    I also selected data that is complex but familiar.
+I have worked in higher education for years, so, while IPEDS data is itself complicated, it helps that I am used to using higher education data, again helping to keep the scope manageable.
 
-    The general backup plan is to simplify the IPEDS dataset down quite a bit.
-We may even be able to get away with using just one domain of IPEDS data if needed.
-Since a major goal of this project is to demonstrate synthesizing IPEDS data, we may also only keep data that looks relatively clean to start with.
-We will also get a sense of which features in the data may be easier to model, again providing a way to simplify the scope to accommodate competing needs.
+    The result of trimming the scope is that I leave more time to play with generative modeling.
+While the final generative modeling notebook is pretty streamlined, it did not start that way.
+I had the opportunity to explore a number of models spanning different algorithms and a range of hyperparameters.
 
-    To wrap up with an explicit acknowledgment of risks for my projects, the biggets worry is that the scope grows to an unruly size, so there are a number of comments in this report speaking to ways to keep that in check, spanning from stricter time scoping to simplifying the data to sticking with easier-to-implement modeling.
+    My initial main challenge that I was looking out for was much higher complexity in the IPEDS data.
+To be fair, the IPEDS data took a good chunk of the time in this project to sort through, and, as stated earlier, I purposely narrowed down the data to try and address this challenge.
+
+    The next challenge was around how much automation to bake into the project.
+Ideally, I wanted to build out a fully automated pipeline, but I ended up finding that the extra effort to implement this would not add to the value of this iteration of the project.
+Instead, I have some placeholders and plans for how to increase automation in future iterations.
+These primarily look like leveraging the orchestration script, adding database objects such as stored procedures and triggers, and building out some utility scripts that help to add in new data.
+For example, a we could built a utility script that reads a new raw data file, pulls the headers, creates an SQL script to create a new table, and push the data into the database, as one example.
+This is in contrast to building all this out manually as we do at the moment, and it is one example of where we could use such a script.
+
+    In terms of ongoing challenges, I am currently working through evaluations of the synthetic data.
+I have refined the process quite a bit, but there are still some remaining areas to smooth out.
+For example, I already had to drop some features due to their negative performance on generative model training.
+I would like to explore ways to tinker with the generative modeling to encompass a broader range of data.
+This could look like changing hyperparameters, but it is more likely that it means curating different modeling datasets to answer specific questions.
+
+    I already leveraged one of my backup plans in that I needed to descope the range of IPEDS data that I originally looked to include. Three separate times.
+
+    If I run into any serious issues with generative modeling, it is an option to test out different libraries.
+I have a few in mind based on other research papers I have worked through in the past, but I am not realistically expecting to switch libraries this far into the project.
+
+    A final backup plan would be what to do if the steps after generative modeling end up a bit too plain.
+For example, while the synthetic data we create now may look good on the surface, it may not be as useful as we would like it to be in terms of regression, classification, or other approach.
+The simpler plan here would be to expand the evaluation steps for the synthetic data and generative modeling since this already includes some extra modeling.
+It could be an option to return to the IPEDS files and select a wider range of features, but I would need to see some serious trouble before wanting to shift back that far in the pipeline to start redoing work.
+
+    I do want to call out that, since this report comes most of the way through the planning, design, and implementation of the project, I am hoping that there are not too many surprises left, if any.
+I have also already taken steps to mitigate my biggest worry of scope creep, so I am not as worried about that creeping back in at this point.
+It will require not letting final modeling take up too much space, but, fortunately, the project feels pretty robust as it is now, so there may be an opportunity to find a way to wrap things up soon.
 

### 2025-03-15 artifacts/report/msds_data_mining_project_report.tex
adjust conclusion section in report

@@
 \section{Conclusion}
+    To recap, we are using three domains of IPEDS data to test out generative modeling to produce synthetic tabular data, preparing for the reality where we have less and less higher education data in the coming years.
+We set up a small sample of a full ETL pipeline, send the IPEDS data through, generate synthetic data, and do some exploratory modeling on that data.
+This approach is probably over-engineered for the scope of this project, but it lays the groundwork for pulling in much more IPEDS data and running more complicated analyses in the future.
 
-    project summary
-    key findings
-    future work
+    At this point, I am still experimenting with the outputs from the generative modeling.
+Most of the pipeline and project are complete, and the third iteration of this report and the associated slides will fill in details about modeling on the the synthetic data itself.
 
-    This approach is probably over-engineered for the scope of this project, but it lays the groundwork for pulling in much more IPEDS data and running more complicated analyses
+    The biggest findings so far have been some of what we expected to see: the IPEDS data contains a lot more complexity than what it looks like at first.
+We address this by descoping the data that we work with in this pass.
+In future iterations, we will incorporate much more IPEDS data.
+That will result in a more robust and complicated warehouse, generative modeling, and exploration of questions with the synthetic data.
+But, for now, the scope here feels healthy for a first iteration on a project of this sort.
 
-    Emphasize where the project currently is at when updating report and slides
+    One area to call out that I do not plan to have an answer for in this project is how we would pivot from IPEDS institutional-level data to student-level data, the latter being the area where are likely to see the biggest drop-off in future years.
+When doing initial planning for this project, I expected that there would be a couple of obvious areas in which to pivot from IPEDS data to student-level data, but those areas have not revealed themselves.
+One issue is that student-level data by its nature of being scoped to individual students is not and should not be in large aggregated and publicly available data such as what we find on IPEDS, primarily for privacy concerns.
+There are many questions that the batch of IPEDS data in this project raise that could lead us to exploring student-level data, but I do not see any reason to go down that path at this point.
 
-    Why synthesize the IPEDS data we use instead of student-level data, the area where we will see a real drop off in coming years? How are aggregates in IPEDS different from student-level data?
+    That all said, while generating synthetic IPEDS data is interesting and valuable, it is likely that generating synthetic student-level data may be the real need in the future.
 
-    There is not much to conclude at the moment.
-We will fill this in as we start doing the data work of the project.
-What I expect to see here is that we trimmed down the IPEDS data, scoped the use case for the project to be a bit narrower, spent more time explore synthetic data and its associated generative modeling, and picked a simpler overall output, likely sticking supervised modeling.
-For a surprise finish, we may end up deciding to pull some data out and run it through a separate business intelligence tool like Tableau with the idea that end users of the production pipeline may want to leverage tools such as that for their purposes, though this would only come into play if we find that the project turns out to be too quick and simple to get through.
-
-    As for future iterations of the project, I expect that the biggest area of grows would be to incorporate more IPEDS data to answer a broader range of questions.
-I do not expect radical changes in the pipeline architecture in next iterations of the project, but I do expect additions and expansions in order to accommodate particular questions of interest inherent in the new data.
+    In terms of the synthetic data itself, while I am still working on that layer of the pipeline and project, so far it appears to work as desired, allowing us to create new data that is similar to the statistical trends we see in the real data.
+I have pulled in some of my higher education data experience when working with this batch of IPEDS data, but a first next step in a future iteration of the project would be to highlight areas where there are known difficulties and confusion with higher education data and exploring how the generative modeling handles those.
+And, as said above, the pipeline and approach here could use thorough testing and enhancement related to incorporating a much wider range of IPEDS data, a more complicated warehouse, and improved scope and usage of generative modeling.
 
@@
 Note that the biggest difference between the list here and the larger summary in data/data\_notes.md in the source repository is that we drop gender and race/ethnicity disaggregates here.
+
+    \subsection{Data from Raw Files}
 

### 2025-03-16 artifacts/report/msds_data_mining_project_report.tex
clean up conclusion section of report

@@
 
-    \subsection{Pipeline Walkthrough}
     You can view the source code at \href{https://github.com/jhleakakos/msds-data-mining-project}{https://github.com/jhleakakos/msds-data-mining-project}.
 
+    \subsection{Pipeline Walkthrough}
         We start by using a Python script as a placeholder for some sort of orchestration system.
@@
     \subsection{IPEDS Data}
-
-        Note: you can find a more detailed explanation of the IPEDS data we use in the appendix, and you can find a more detailed summary of the IPEDS domains we work with in data/data\_notes.md in the GitHub repository.
+        Note: you can find a file and field listing of the IPEDS data we use in the appendix, and you can find a more detailed summary of the IPEDS domains we work with in data/data\_notes.md in the GitHub repository.
 
@@
         \end{itemize}
+
+        We choose these three domains since they give us a basic count of students coming into higher education institutions each year and the number of completions --- a proxy for a type of output --- within that same year.
 
@@
 
-    \subsubsection{Data Analysis}
-        Will fill this section in more for the next pass of the report.
+    \subsubsection{Modeling and Analysis}
+        Since this project is focused on generating synthetic data, we end up shifting some of the scope that would be given to typical analysis to focus more on evaluating the generative modeling and its synthetic data output.
+I do have a list of questions related to feature importance in terms of enrollment and completion numbers as well as a range of classification and clustering experiments that I planned on testing out.
+But, the generative modeling, while easy enough in a technical sense using the libraries we take advantage of, ended up taking a lot longer in terms of understanding what was going on and how to responsibly impelement it in the pipeline.
 
-        This is the remaining chunk of work that I have to complete at this point.
-The current thinking is that we will lean into testing out different modeling or mining techniques across the real data, across the synthetic data, and across both combined together.
-The focus of this project is on synthesizing IPEDS data, so we need to explore how incorporating the synthetic data affects common output tasks that we will want to run on different subsets of IPEDS data.
-
-        Note that this will end up looking a lot like the machine learning score evaluation technique that we explained earlier.
-This means our modeling analysis will blur the line with evaluation for this project, but it will help us to gain more fluency in working with synthetic data generally and the use of the specific libraries we use with IPEDS data.
+        In terms of generative modeling, we end up generating a smaller synthetic dataset that is focused on a handful of categorical features related to institutional characteristics along with matching enrollment and completion numbers.
+I will raise some questions about this approach later in the discussion section, but, for this pass of the project, it allows us to double the amount of data we have in the real dataset, and, as we will see in the evaluation section, the synthetic data is sufficiently similar to the real data to either substitute for or supplement that real data.
 
@@
 
+        Once we finish up generative modeling, we do test out linear regression to help evaluate the quality of the synthetic data.
+We start by confirming that there is enough of a linear relationship between the inputs and outputs to continue using a linear model.
+We then pull out a test set from the real data to check the trained linear models.
+
+        Next, we have two sets of regressions focusing on enrollment and completions.
+We split each of these two into three different training sets: the real data, the synthetic data, and a combination of the real and synthetic data.
+This leaves us with six total models to evaluate.
+We use mean squared error (MSE) and adjusted r-squared ($R^2_a$) to evaluate each model.
+The focus here is on checking if we see similar values for each of these metrics across all three training sets for a given model.
+If we do see this, then we have some evidence that the synthetic data is resulting in similar evaluative outputs as the real data, either on its own or in combination with the real data.
+
 \section{Evaluation}
+
+    GAN losses
+    Pick a feature or two and include the image in here
+    Pick a feature pair or two and include the image in here
+    100% score for diagnostic report, confirming the data is valid (explain valid)
+
+    evaluate quality gives 87.1%
+    87.2% score for column shapes
+    87.0% score for column pair trends
+    explain column shapes and column pair trends
+
+    quality and diagnostic reports
+    GAN losses
+    Pick a couple of images to display
+
+    Which column pair trends are most and least valuable in terms of score
+    Use visual to help identify these
+
+    Include main column shapes image that shows score for all features
+
+    TVComplement and KSComplement
+    TV is total variation distance (TVD) compares the probability of each category in the real and synthetic dataset and compares them
+    KS is Kolmogorov-Smirnov statistic that converts each numeric feature into a cumulative distribution function and finds the maximum distance between the CDFs
+    https://docs.sdv.dev/sdmetrics/metrics/metrics-glossary/tvcomplement
+    https://docs.sdv.dev/sdmetrics/metrics/metrics-glossary/kscomplement
+
+    StatisticSimilarity
+    Only for numeric columns
+    Can compare mean, median, or std between real and synthetic matching features
+    Measures the absolute value of the difference between the real and synthetic feature for the metric scaled to the range of the real feature
+    Can run on the two numeric target features
+    https://docs.sdv.dev/sdmetrics/metrics/metrics-glossary/statisticsimilarity
+
+    Skewed numeric features
+
+    Going back, we should probably train the synthetic data on a portion of the real data so that we have a completely held-out test set
+
+    Include modeling stats -- the three models are close to each other for each target feature
+    Check MSE and r-squared for each model for enrollments and completions
+    Explain what we are evaluating here
+
     The first evaluations will capture how similar the synthetic data is to the real data.
@@
 \section{Discussion}
+
+    Do we generate one larger dataset and then push that into the warehouse?
+Do we generate a bunch of smaller datasets knowing that the cross-section between them earlier in the pipeline may get lost?
+    Appropriate and responsible ways to use generative modeling?
+    Could an individual institution take this work and use it in tandem with their internal student-level data to get a personalized picture?
+
+    In discussion section, emphasize that we did spend a lot more time on generative modeling and cut scope after to leave space for generative modeling exploration
+    In discussion, include what lessons we learned in this project
+
     My goal for this project is to stick within a 3-4 week limit, if possible.
@@
 This approach is probably over-engineered for the scope of this project, but it lays the groundwork for pulling in much more IPEDS data and running more complicated analyses in the future.
+I also leaned into the architecting portion of the project, focusing on system design rather than picking one area to prioritize above the rest, though we do lean into generative modeling as the new-ish piece that we incorporate here.
 
-    At this point, I am still experimenting with the outputs from the generative modeling.
-Most of the pipeline and project are complete, and the third iteration of this report and the associated slides will fill in details about modeling on the the synthetic data itself.
+    Overall, we have a full pipeline with hooks in it for future improvement, successful implementation of generative modeling along with evaluative metrics that support that claim, the ability to increase the scope of IPEDS data we work with, and a range of other questions and enhancements we can dig into after this iteration of the project.
+Incorporating generative modeling did add a fresh aspect for me into this work, and I am glad I did it, though it meant that I had to descope other areas to accommodate the new modeling steps.
 
-    The biggest findings so far have been some of what we expected to see: the IPEDS data contains a lot more complexity than what it looks like at first.
+    The biggest findings have been some of what we expected to see: the IPEDS data contains a lot more complexity than what it looks like at first.
 We address this by descoping the data that we work with in this pass.
@@
 But, for now, the scope here feels healthy for a first iteration on a project of this sort.
+If I could go back and scope this project again, I would consider building two smaller projects and then combining them --- one project focused on IPEDS data and the other focused on generative modeling.
 
-    One area to call out that I do not plan to have an answer for in this project is how we would pivot from IPEDS institutional-level data to student-level data, the latter being the area where are likely to see the biggest drop-off in future years.
+    One area to call out that I do not plan to have an answer for in this iteration of the project is how we would pivot from IPEDS institutional-level data to student-level data, the latter being the area where are likely to see the biggest drop-off in future years.
 When doing initial planning for this project, I expected that there would be a couple of obvious areas in which to pivot from IPEDS data to student-level data, but those areas have not revealed themselves.
 One issue is that student-level data by its nature of being scoped to individual students is not and should not be in large aggregated and publicly available data such as what we find on IPEDS, primarily for privacy concerns.
-There are many questions that the batch of IPEDS data in this project raise that could lead us to exploring student-level data, but I do not see any reason to go down that path at this point.
 
-    That all said, while generating synthetic IPEDS data is interesting and valuable, it is likely that generating synthetic student-level data may be the real need in the future.
+    Probably the best path forward to include student-level data would be to modify the pipeline to prepare IPEDS data in such a way that individual institutions can clone the pipeline and incorporate their specific student-level data as needed.
+This idea has actually become my main area of interest for next steps after completing this project.
+It would require building out more utility functionality that lets users customize specifics in the pipeline.
+For instance, we could build a series of scripts that create separate SQL scripts or Python EDA scripts to incorporate new data, something akin to building a format for configuration that the system could be responsive to.
+But that would pivot us quite a bit away from the focus on generative modeling here, so we will leave it for a future iteration.
 
-    In terms of the synthetic data itself, while I am still working on that layer of the pipeline and project, so far it appears to work as desired, allowing us to create new data that is similar to the statistical trends we see in the real data.
+    In terms of the synthetic data itself, so far it appears to work as desired, allowing us to work with new data that is similar to the statistical trends we see in the real data.
 I have pulled in some of my higher education data experience when working with this batch of IPEDS data, but a first next step in a future iteration of the project would be to highlight areas where there are known difficulties and confusion with higher education data and exploring how the generative modeling handles those.

### 2025-03-17 artifacts/report/msds_data_mining_project_report.tex
update evaluation section of report

@@
 \section{Evaluation}
+    Due to scope and time, the evaluations in this iteration of the project focus on determining how good our synthetic data is.
+The original intention was to do through this and then do a full extra round of analysis on top of the data, including the synthetic data, to glean further insights, but we instead give more time and space to generative modeling and evaulation of those models and their outputs, sticking closer to the theme of the project.
 
-    GAN losses
-    Pick a feature or two and include the image in here
-    Pick a feature pair or two and include the image in here
-    100% score for diagnostic report, confirming the data is valid (explain valid)
+    We leverage evaluation metrics in the Synthetic Data Vault Project to evaluate the synthetic data, including determining how similar the synthetic data is to the real data.
+Earlier we spoke of per-attribute statistics and machine learning score.
+These two categories capture most of what we use to evaluate the synthetic data.
 
-    evaluate quality gives 87.1%
-    87.2% score for column shapes
-    87.0% score for column pair trends
-    explain column shapes and column pair trends
+    All of the metrics here aside from machine learning score come from SDMetrics, a library in the Synthetic Data Vault Project.
+The summaries of how the metrics here work are informed by the SDMetrics documentation.
 
-    quality and diagnostic reports
-    GAN losses
-    Pick a couple of images to display
+    We start by checking out the GAN losses when training the CTGAN model.
 
-    Which column pair trends are most and least valuable in terms of score
-    Use visual to help identify these
+    \begin{figure}[h]
+        \centering
+        \caption{GAN losses while training CTGAN model}
+        \Description{Line plot of generator and discriminator losses during CTGAN training.}
+        \includegraphics[width=\linewidth]{gan_losses}
+    \end{figure}
 
-    Include main column shapes image that shows score for all features
+    We see expected progression for the generator and discriminator loss.
+The discriminator loss stays relatively flat while the generator loss decreases.
+GAN training is notoriously difficulty, including for our use case, so it is promising that we see expected losses in a relatively short time.
+The final CTGAN we train finishes training in under a minute.
 
-    TVComplement and KSComplement
-    TV is total variation distance (TVD) compares the probability of each category in the real and synthetic dataset and compares them
-    KS is Kolmogorov-Smirnov statistic that converts each numeric feature into a cumulative distribution function and finds the maximum distance between the CDFs
-    https://docs.sdv.dev/sdmetrics/metrics/metrics-glossary/tvcomplement
-    https://docs.sdv.dev/sdmetrics/metrics/metrics-glossary/kscomplement
+    SDMetrics includes a diagnostic report that checks the validity of the synthetic data.
+This checks that primary key fields are unique, that numeric and datetime fields fall within the minimum and maximum of the range for that field in the real data, and that categorical features contain the same categories as in the real data (\href{https://docs.sdv.dev/sdmetrics/reports/diagnostic-report/whats-included}{https://docs.sdv.dev/sdmetrics/reports/diagnostic-report/whats-included}).
+The diagnostic report returns 100\%, indicating that our synthetic data is valid according to these conditions.
 
-    StatisticSimilarity
-    Only for numeric columns
-    Can compare mean, median, or std between real and synthetic matching features
-    Measures the absolute value of the difference between the real and synthetic feature for the metric scaled to the range of the real feature
-    Can run on the two numeric target features
-    https://docs.sdv.dev/sdmetrics/metrics/metrics-glossary/statisticsimilarity
+    Next we check out the SDMetrics quality report.
+This report checks how well the synthetic data matches the statistical properties found in the real data.
+These include the per-attribute comparisons we spoke about earlier.
+Per-attribute comparisons check the same feature in the real and synthetic datasets, calculate metrics such as the mean, median, or standard deviation, and measure how similar these are between the dataset.
 
-    Skewed numeric features
+    \begin{table}
+        \caption{SDMetrics Quality Report}
+        \begin{tabular}{ll}
+            \toprule
+            Property & Score\\
+            \midrule
+            Column Shapes & 0.872 \\
+            Column Pair Trends & 0.870 \\
+            Overall Score & 0.871 \\
+            \bottomrule
+        \end{tabular}
+    \end{table}
 
-    Going back, we should probably train the synthetic data on a portion of the real data so that we have a completely held-out test set
+    The quality report includes two primary metrics: column shapes and column pair trends (table 1).
+Column shapes performs the per-attribute metric from above.
+For categorical features, column shapes compares the probability of each category in a given feature in the real and synthetic data and compares them (\href{https://docs.sdv.dev/sdmetrics/metrics/metrics-glossary/tvcomplement}{https://docs.sdv.dev/sdmetrics/metrics/metrics-glossary/tvcomplement}).
+For numeric features, column shapes compares the cumulative distribution function (CDF) for the feature in the real and synthetic data and finds the maximum distance between them (\href{https://docs.sdv.dev/sdmetrics/metrics/metrics-glossary/kscomplement}{https://docs.sdv.dev/sdmetrics/metrics/metrics-glossary/kscomplement}).
+With the way these are calculated, higher numbers are better, and scores closer to 100\% mean the synthetic data is more statistically similar to the real data.
 
-    Include modeling stats -- the three models are close to each other for each target feature
-    Check MSE and r-squared for each model for enrollments and completions
-    Explain what we are evaluating here
+    \begin{table}
+        \caption{SDMetrics Column Shapes}
+        \begin{tabular}{lll}
+            \toprule
+            Column & Metric & Score \\
+            \midrule
+            year & TVComplement & 1.000 \\
+            state\_abbr & TVComplement & 0.882 \\
+            bea\_region & TVComplement & 0.897 \\
+            highest\_level & TVComplement & 0.938 \\
+            is\_degree\_offering & TVComplement & 0.945 \\
+            is\_hbcu & TVComplement & 0.973 \\
+            is\_tribal\_institution & TVComplement & 0.974 \\
+            geographic\_status & TVComplement & 0.892 \\
+            date\_closed & KSComplement & 0.167 \\
+            institutional\_category & TVComplement & 0.945 \\
+            control\_affiliation & TVComplement & 0.906 \\
+            enrollment & KSComplement & 0.887 \\
+            completions\_number\_students & KSComplement & 0.932 \\
+            \bottomrule
+        \end{tabular}
+    \end{table}
 
-    The first evaluations will capture how similar the synthetic data is to the real data.
-We will explore the evaluation metrics in the Synthetic Data Vault Project, and, on top of those metrics, a couple of extra candidates that stand out are per-attribute comparisons and machine learning score.
-Per-attribute comparisons check the same feature in the real and synthetic datasets, calculate metrics such as mean or standard deviation, and determine the quality of the synthetic data based on how close these measures are across the datasets.
-Machine learning score trains classifiers on the real and synthetic datasets and compares the performance of those models.
-For example, if we use both classifiers to check something like accuracy or F1 score on a held-out test set, we can evaluate how similar the synthetic data is to the real data based on comparing those scores.
+    \begin{figure}[h]
+        \centering
+        \caption{Column Shapes Scores}
+        \Description{Bar plots of SDMetrics column shapes scores}
+        \includegraphics[width=\linewidth]{column_shapes}
+    \end{figure}
 
-    We can also look at more advanced evaluation metrics.
-One that has stood out to me predicting each feature in a dataset with all of the other features.
-If we find the predictive power in the real and synthetic datasets are similar, then, like above, that can give us a sense of how similar the datasets themselves are.
+    Looking at table 2 and figure 2, we see some outliers.
+year gives perfect statistical similarity since every row has the same value at this point.
+Later iterations can load more than one year of data, so we want to keep this field around.
+date\_closed is mostly null, so that is likely why this field is so off --- there is not enough information in the real data for the model to learn meaningful patterns.
+The remaining features show more similarity between the real and syntethetic features indicated by their scores above 0.5.
+We do not want scores to be too close to 1.0 since that may result in a situation where the synthetic data is too real and can reveal confidential information from the real data.
+We end up dropping year and date\_closed for later modeling, so we are not worried about these features for now, and the remaining features look good.
 
-    There are any number of other evaluative metrics for synthetic data.
-We will explore what is available with the Synthetic Data Vault Project that we use and will clarify here as we learn more during experimentation.
-It is worth calling out here that we could expand the evaluation of the synthetic data itself, taking up more scope from later steps such as regression or classification or such.
-The risk of not spending enough time evaluating the synthetic data is that we may end up testing out further modeling on an ill-fitting synthetic dataset.
+    \begin{figure}[h]
+        \centering
+        \caption{Column Shapes Plot for bea\_region}
+        \Description{Bar plot of SDMetrics column shape for bea\_region}
+        \includegraphics[width=\linewidth]{column_plot_bea_region}
+    \end{figure}
 
-    At this stage, we have run a pass of generative modeling and added summary statistics and visualizations in the generative modeling notebook that you can explore in the source code.
-We also drop two features that we originally meant to include due to their negative effect on generative modeling.
-I will hold off on doing a more detailed summary until the next pass of this report, though the initial metrics look promising, with the summary numbers falling into ranges that hint that the synthetic data is picking up on patterns in the real data without mirroring the real data so closely that we run into privacy concerns.
+    \begin{figure}[h]
+        \centering
+        \caption{Column Shapes Plot for enrollment}
+        \Description{Line plot of SDMetrics column shape for enrollment}
+        \includegraphics[width=\linewidth]{column_plot_enrollment}
+    \end{figure}
 
-    Moving on from synthetic data, my main intention is to build regression models on total enrollment and completion numbers per institution.
-I am curious to see which descriptive features help the most in predicting enrollments and completions.
-This would be a supervised regression task that we have built the warehouse and modeling dataset around.
-We can use metrics like mean squared error (MSE) or adjusted r-squared ($R^2_a$) to measure these regression models.
-These evaluations give us firm numbers we can use to determine how well our modeling is doing.
-We can compare across the three different groups of input data to see how well the synthetic data is doing, and we can focus on the modeling for the real data to see if we can glean insights out of the IPEDS data.
+    SDMetrics also evaluates column pairs between the real and synthetic data.
+These are not as informative for our current focus, though we would want to return to these when evaluating machine learning score, a metric we explain next.
+But first, here are two sample column pair plots.
 
-    I am not sure yet if we will include unsupervised learning techniques.
-Admittedly, the outputs for unsupervised learning can be more difficult to work with.
-Clustering is the first unsupervised modeling that comes to mind.
-I often think of clustering as a much more advanced version of EDA.
-We are not really sure what patterns exist in the data, so we supercharge standard EDA and see what patterns we uncover when leveraging machine learning.
-But we do not have a clean evaluation metric with clustering.
-We will need to interactively explore the clustering outputs, at least at first.
-At some point, we can determine a set number of clusters and pivot the clustering into a classification approach, but the more interesting piece here for me is the clustering itself.
+    \begin{figure}[h]
+        \centering
+        \caption{Column Pair Plot for enrollment and completions}
+        \Description{Scatter plot of SDMetrics column pair relationship for enrollment and completions}
+        \includegraphics[width=\linewidth]{column_pair_plot_enrollment_completion}
+    \end{figure}
 
-    While I feel like there are loads of evaluative methods to explore, I am also interested in looking at some of the algorithms from the second data mining course.
-I have leveraged things like similarity scores in other projects, but I have not experimented with, for instance, FP growth, and it would be interesting to possibly pick a brand new algorithm to plop at the end of the pipeline and figure out what it means to evaluate that algorithm and what it reveals to us.
-But, similar to with unsupervised learning, my desire to explore synthetic data may mean leaving out some of the new algorithms from the second course for now.
-This would be an area to explore more in the next iteration of the project though.
+    \begin{figure}[h]
+        \centering
+        \caption{Column Pair Plot for is\_degree\_offering and completions}
+        \Description{Box plot of SDMetrics column pair relationship for is\_degree\_offering and completions}
+        \includegraphics[width=\linewidth]{column_pair_plot_is_degree_offering_completion}
+    \end{figure}
+
+    Finishing out per-attribute metrics, SDMetrics can compute the statistical similarity of the mean, median, and standard deviation of a feature across the real and synthetic data, measure the absolute value of the difference between one of these metrics across the datasets and then scaled by the range of the real data for that feature (\href{https://docs.sdv.dev/sdmetrics/metrics/metrics-glossary/statisticsimilarity}{https://docs.sdv.dev/sdmetrics/metrics/metrics-glossary/statisticsimilarity}).
+
+    Next we finish up with machine learning score, training regressors on the real data, synthetic data, and a combination of both and comparing the performance of them with one another.
+If we find the MSE or $R^2_a$ is similar between the three training sets, then we have some evidence that the synthetic data is similar enough to the real data to stand in for it or to be combined with it without reducing model performance.
+
+    \begin{table}
+        \caption{Machine Learning Score}
+        \begin{tabular}{llll}
+            \toprule
+            Metric & Training Set & Target & Value \\
+            \midrule
+            MSE & Real & Enrollment & 101,200,325.55 \\
+            MSE & Synthetic & Enrollment & 111,006,681.53 \\
+            MSE & Combined & Enrollment & 101,978,162.27 \\
+            $R^2_a$ & Real & Enrollment & 0.19 \\
+            $R^2_a$ & Synthetic & Enrollment & 0.11 \\
+            $R^2_a$ & Combined & Enrollment & 0.19 \\
+            MSE & Real & Completions & 116,066,105.90 \\
+            MSE & Synthetic & Completions & 133,327,873.83 \\
+            MSE & Combined & Completions & 129,922,536.22 \\
+            $R^2_a$ & Real & Completions & -0.01 \\
+            $R^2_a$ & Synthetic & Completions & -0.07 \\
+            $R^2_a$ & Combined & Completions & -0.04 \\
+            \bottomrule
+        \end{tabular}
+    \end{table}
+
+    The values for each of the machine learning score metrics are pretty similar, hinting that the synthetic data is of pretty good quality.
+These are not necessarily the best linear regression models in terms of predicting enrollment or completion numbers, but the fact that the MSE and $R^2_a$ values are similar across the real, synthetic, and combined datasets means the generative modeling did its job.
+
+    We do have some skewed numeric features and imbalanced categorical features in the real data.
+CTGAN has techniques to handle these, and it looks like those were able to handle these aspects of the real data without trouble.
+We also dropped some features with a large number of categories since they were slowing down generative modeling without providing much upside.
+
+    In future iterations, I would like to include a couple of changes.
+First is adding a new evaluation metric where we predict each feature in a dataset with all the other features, repeating this across the real, synthetic, and combined datasets.
+If we find the predictive power in all three datasets are similar, then that can give us further evidence that the generative modeling is producing high-quality synthetic data.
+
+    At this stage, we have run successful generative modeling on our subset of IPEDS data.
+These metrics look promising, with the summary numbers falling into ranges that hint that the synthetic data is picking up on patterns in the real data without mirroring the real data so closely that we run into privacy concerns.
+
+    I chose to descope any further modeling beyond these in order to keep the scope from ballooning.
+Experimenting with and evaluating the generative models and their synthetic data outputs took longer than expected, though the time was well spent.
+The generative modeling base that we have in this iteration of the project will serve us well when we look to expand use cases in the future.
 
@@
     In discussion, include what lessons we learned in this project
+
+    Should we train generative models on smaller or combined datasets?
+Do we lose anything by training them on smaller datasets later in the pipeline?
+Depends on use case.
+If we want the warehouse to have synthetic data, then we should model earlier.
+If we are using synthetic data only later for analysis, then maybe it doesn't matter as much.
 

### 2025-03-18 artifacts/report/msds_data_mining_project_report.tex
update conclusion section of report

@@
 \section{Discussion}
-
-    Do we generate one larger dataset and then push that into the warehouse?
-Do we generate a bunch of smaller datasets knowing that the cross-section between them earlier in the pipeline may get lost?
-    Appropriate and responsible ways to use generative modeling?
-    Could an individual institution take this work and use it in tandem with their internal student-level data to get a personalized picture?
-
-    In discussion section, emphasize that we did spend a lot more time on generative modeling and cut scope after to leave space for generative modeling exploration
-    In discussion, include what lessons we learned in this project
-
-    Should we train generative models on smaller or combined datasets?
-Do we lose anything by training them on smaller datasets later in the pipeline?
-Depends on use case.
-If we want the warehouse to have synthetic data, then we should model earlier.
-If we are using synthetic data only later for analysis, then maybe it doesn't matter as much.
-
-    My goal for this project is to stick within a 3-4 week limit, if possible.
+    My goal for this project was to stick within a 3-4 week limit, and I was able to stick within this timeframe.
 I tend to drag out projects, taking the opportunity to add increased complexity and explore new items I am curious about.
@@
 
-    The first area to keep an eye on scope is with the data used in the project.
-We start by selecting three IPEDS domains among many.
+    The first area in which to keep an eye on scope is with the data used in the project.
+We start by selecting three IPEDS domains out of many.
 We then systematically trim the datasets down in stages, as described earlier.
@@
 Two of those are numeric targets we can use for regression.
-We could also put some boundary cutoffs for those fields to aid with classification.
+We could also put some boundary cutoffs for those fields to test out classification.
 We have a couple of more likely features in terms of explanatory power, and we have a few extras to test out in terms of importance.
@@
 
+    I also selected data that is complex but familiar.
+I have worked in higher education for years, so, while IPEDS data is itself complicated, it helps that I am used to using higher education data, again helping to keep the scope manageable.
+There are a lot of interesting higher education questions that got dropped while trimming the data down, but it will be feasible to add those back in in future iterations based on the design of the system.
+
+    So, I ended up leveraging one of my backup plans in that I needed to descope the range of IPEDS included in the project. Three separate times.
+
     Next is the use of the database.
 This is a personal thing for me, but I am more familiar working in databases over the years, so doing work in the database simplifies the amount of time needed for me to set up a production-esque system.
+We also run into databases everything in real-world systems, so a project that mirrors a production pipeline, even if on a very small scale, must have database tech included in it, in my experience.
 
-    I also selected data that is complex but familiar.
-I have worked in higher education for years, so, while IPEDS data is itself complicated, it helps that I am used to using higher education data, again helping to keep the scope manageable.
+    Finally, a big area of scope reduction is at the end of the pipeline.
+I originally intended to implement both generative modeling and analytical modeling.
+Generative modeling is key to the focus of this project, but I was unclear if the project would feel incomplete if it did not have extra layers of analytical modeling as well.
+It was not until this final update for the project and deliverables that I decided to stick with the generative modeling and leave the analytical modeling until a future expansion of the project.
+Since I decided to explore generative modeling early in the planning stages, that took priority over analytical modeling, and the generative modeling itself took a significant chunk of time to familiarize myself with, experiment with, and implement.
+To add extra analytical modeling at this stage feels like it would balloon the scope and pull the focus of the project elsewhere.
 
-    The result of trimming the scope is that I leave more time to play with generative modeling.
-While the final generative modeling notebook is pretty streamlined, it did not start that way.
-I had the opportunity to explore a number of models spanning different algorithms and a range of hyperparameters.
+    Speaking of generative modeling, even after all the tinkering I did for this project, I still have some lingering questions.
+The first is where the generative modeling should happen.
+The approach in this project was to pull the analysis dataset out of the warehouse and synthesize more data mimicking that dataset.
+If we zoom out and think about a pipeline and warehouse that needs to be more general purpose, there is a case to synthesize data earlier, possibly even keeping real and synthetic fact and dimension tables in the warehouse.
+If the goal of the synthetic data is to distribrute a final dataset, then our approach here works well.
+If the goal is to incorporate synthetic data as a core piece of the warehouse, then we may want to synthesize data earlier.
 
-    My initial main challenge that I was looking out for was much higher complexity in the IPEDS data.
+    I am still on the fence about the right answer here.
+I think the location of generative modeling in this pass of the project is correct, but the big downside is that we are generating new data based on a very small and bespoke subset of the overall IPEDS data that we start with in this project, and that is already a big reduction in terms of the overall IPEDS data that is available across more domains.
+With my current understanding of generative modeling, we may want the modeling to be able to look at a wider range of IPEDS data when learning patterns to synthesize since we may drop some important relationships between features before we hit generative modeling in this project.
+We could generate one much larger dataset and then push that into the warehouse, possibly adding generative modeling as a step between the EDA after the data lake and before transforming data into the dimensional model.
+The trade-off though is that it feels strange to add generative modeling in too early, like we are building a warehouse and pipeline primarily to serve synthetic data, not real data.
+This is one of my biggest questions that is left at the end of this iteration of the project.
+
+    Another design question that has become more prominent as the project progressed has been how individual institutions may be able to combine their student-level data with the system here.
+As stated earlier, while synthesizing institutional-level data is valuable, it is likely that synthesizing student-level data is where the bigger value lies.
+One approach I can think of is to modify the pipeline to read in common student-level data, model it in the data lake and warehouse, and then prepare for it in terms of generative modeling, wherever we put that modeling step.
+I do not want to derail the report here with a lot of theorizing about how to incorporate student-level data, but it is worth noting that this is, along with where to plop the generative modeling, the biggest question I am left with at the end of the project.
+
+    The initial main challenge that I was looking out for was much higher complexity in the IPEDS data.
 To be fair, the IPEDS data took a good chunk of the time in this project to sort through, and, as stated earlier, I purposely narrowed down the data to try and address this challenge.
@@
 
-    In terms of ongoing challenges, I am currently working through evaluations of the synthetic data.
-I have refined the process quite a bit, but there are still some remaining areas to smooth out.
-For example, I already had to drop some features due to their negative performance on generative model training.
-I would like to explore ways to tinker with the generative modeling to encompass a broader range of data.
-This could look like changing hyperparameters, but it is more likely that it means curating different modeling datasets to answer specific questions.
+    In terms of ongoing challenges, I think the big one is keeping an eye on how to evaluate the synthetic data, especially if we start modifying the pipeline as described throughout this section.
+While I am comfortable with running evaluation metrics, interpreting them across different use cases will take a lot more experience.
+It would be good to put processes in place to track evaluation metrics over time and for different uses in order to improve the value and quality of the synthetic data.
+For example, we do not want synthetic data to 100\% match the real data, but we also do not want it to be so different that the conclusions you reach with the synthetic data are invalid.
 
-    I already leveraged one of my backup plans in that I needed to descope the range of IPEDS data that I originally looked to include. Three separate times.
-
-    If I run into any serious issues with generative modeling, it is an option to test out different libraries.
-I have a few in mind based on other research papers I have worked through in the past, but I am not realistically expecting to switch libraries this far into the project.
-
-    A final backup plan would be what to do if the steps after generative modeling end up a bit too plain.
-For example, while the synthetic data we create now may look good on the surface, it may not be as useful as we would like it to be in terms of regression, classification, or other approach.
-The simpler plan here would be to expand the evaluation steps for the synthetic data and generative modeling since this already includes some extra modeling.
-It could be an option to return to the IPEDS files and select a wider range of features, but I would need to see some serious trouble before wanting to shift back that far in the pipeline to start redoing work.
-
-    I do want to call out that, since this report comes most of the way through the planning, design, and implementation of the project, I am hoping that there are not too many surprises left, if any.
-I have also already taken steps to mitigate my biggest worry of scope creep, so I am not as worried about that creeping back in at this point.
-It will require not letting final modeling take up too much space, but, fortunately, the project feels pretty robust as it is now, so there may be an opportunity to find a way to wrap things up soon.
+    All-in-all, I am left with a number of questions and concerns at the end of this iteration of the project, a good sign that we accomplished some good work while also having room to enhance and expand the system going forward.
 

### 2025-03-18 artifacts/report/msds_data_mining_project_report.tex
revisions to final report

@@
 Faculty and staff members filled the needs for supporting increasing student enrollments.
-But there was an issue hiding behind all of this growth: what happens when student enrollments systematically fall?
+But there was an issue hiding behind all of this growth: what happens when student enrollments systematically begin to drop?
 
@@
 Depending on who you listen to, this ranges from a demographic cliff ushering in an emergency in higher education to a slower decline in higher education enrollments that will play out over years.
-It is unclear what effect this will have on the near future of higher education.
+It is not clear what effect this will have on the near future of higher education.
 One expected consequence of these declines is that we will see smaller amounts of higher education data relative to previous years.
@@
 We use Python scripts to simulate pulling data from application programming interfaces (APIs), run databases in Docker --- including a very small scale warehouse --- to stand in for real-world database usage, and leverage Jupyter Notebooks for exploratory data analysis (EDA) and modeling, including our core focus of generative modeling to create synthetic tabular data.
-We hold off on saving the final tuned versions of models and putting them into automated pipelines, but that would be an extension that we would include in a production system.
+We hold off on saving the final tuned of models and integrating them into automated pipelines as a future enhancement.
 The main idea is to have all the pieces that we would normally tap into in a regular-enough end-to-end data system to provide a foundation from which to expand on later.
@@
 We run a PostgreSQL database locally in Docker.
-Because Postgres connections are database-specific, we use one shared databse for this layer of the data lake and the coming data warehouse layer, using schemas to separate out the data lake from the data warehouse.
+Because Postgres connections are database-specific, we use one shared database for this layer of the data lake and the coming data warehouse layer, using schemas to separate out the data lake from the data warehouse.
 This allows us to use structured query language (SQL) scripts to move data between the data lake and data warehouse without needing to use any other tooling, simplifying the handling of data in the database.
@@
 This leaves us with six total models to evaluate.
-We use mean squared error (MSE) and adjusted r-squared ($R^2_a$) to evaluate each model.
+We use mean squared error (MSE) and adjusted r-squared ($R^2$) to evaluate each model.
 The focus here is on checking if we see similar values for each of these metrics across all three training sets for a given model.
@@
 The discriminator loss stays relatively flat while the generator loss decreases.
-GAN training is notoriously difficulty, including for our use case, so it is promising that we see expected losses in a relatively short time.
+GAN training is notoriously difficult, including for our use case, so it is promising that we see expected losses in a relatively short time.
 The final CTGAN we train finishes training in under a minute.
@@
     Next we finish up with machine learning score, training regressors on the real data, synthetic data, and a combination of both and comparing the performance of them with one another.
-If we find the MSE or $R^2_a$ is similar between the three training sets, then we have some evidence that the synthetic data is similar enough to the real data to stand in for it or to be combined with it without reducing model performance.
+If we find the MSE or $R^2$ is similar between the three training sets, then we have some evidence that the synthetic data is similar enough to the real data to stand in for it or to be combined with it without reducing model performance.
 
@@
             MSE & Combined & Enrollment & 101,978,162.27 \\
-            $R^2_a$ & Real & Enrollment & 0.19 \\
-            $R^2_a$ & Synthetic & Enrollment & 0.11 \\
-            $R^2_a$ & Combined & Enrollment & 0.19 \\
+            $R^2$ & Real & Enrollment & 0.19 \\
+            $R^2$ & Synthetic & Enrollment & 0.11 \\
+            $R^2$ & Combined & Enrollment & 0.19 \\
             MSE & Real & Completions & 116,066,105.90 \\
@@
             MSE & Combined & Completions & 129,922,536.22 \\
-            $R^2_a$ & Real & Completions & -0.01 \\
-            $R^2_a$ & Synthetic & Completions & -0.07 \\
-            $R^2_a$ & Combined & Completions & -0.04 \\
+            $R^2$ & Real & Completions & -0.01 \\
+            $R^2$ & Synthetic & Completions & -0.07 \\
+            $R^2$ & Combined & Completions & -0.04 \\
             \bottomrule
@@
 
-    The values for each of the machine learning score metrics are pretty similar, hinting that the synthetic data is of pretty good quality.
-These are not necessarily the best linear regression models in terms of predicting enrollment or completion numbers, but the fact that the MSE and $R^2_a$ values are similar across the real, synthetic, and combined datasets means the generative modeling did its job.
+    The values for each of the machine learning score metrics are similar, hinting that the synthetic data is of sufficient quality.
+These are not necessarily the best linear regression models in terms of predicting enrollment or completion numbers, but the fact that the MSE and $R^2$ values are similar across the real, synthetic, and combined datasets means the generative modeling did its job.
 
@@
 It was not until this final update for the project and deliverables that I decided to stick with the generative modeling and leave the analytical modeling until a future expansion of the project.
-Since I decided to explore generative modeling early in the planning stages, that took priority over analytical modeling, and the generative modeling itself took a significant chunk of time to familiarize myself with, experiment with, and implement.
+Since I decided to explore generative modeling early in the planning stages, that took priority over analytical modeling, and the generative modeling itself took a significant chunk of time to learn, experiment with, and implement.
 To add extra analytical modeling at this stage feels like it would balloon the scope and pull the focus of the project elsewhere.
@@
 
-    Overall, we have a full pipeline with hooks in it for future improvement, successful implementation of generative modeling along with evaluative metrics that support that claim, the ability to increase the scope of IPEDS data we work with, and a range of other questions and enhancements we can dig into after this iteration of the project.
+    Overall, we have a full pipeline with hooks in it for future improvement.
+We successfully implemented generative modeling along with evaluative metrics that support that show good performance.
+We can relatively easily increase the scope of IPEDS data we work with, at least in a technical sense.
 Incorporating generative modeling did add a fresh aspect for me into this work, and I am glad I did it, though it meant that I had to descope other areas to accommodate the new modeling steps.
